---
title: "Analysis_Metadata"
author: "Lucy"
date: "2023-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load the RDS file
```{r}
# Read the RDS file with full file path
dat_de_master_endo <- readRDS("C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Data/Proteomic_Analysis/dat_de_master_endo.rds")
```


Summary statistics
```{r}
## library to be used 
library(dplyr) 
unique(dat_de_master_endo$Cluster)

# Convert the Cluster column to character
dat_de_master_endo$Cluster <- as.character(dat_de_master_endo$Cluster)

# Drop rows where Cluster is "Endotype B" and save the result in subset_data
subset_data <- subset(dat_de_master_endo, Cluster %in% c("Endotype A", "Endotype C"))
 
                           # OR 

## To calculate the differences between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype B"
subset_data <- dat_de_master_endo %>%
  filter(Cluster != "Endotype B")

                           # OR

subset_data <- dat_de_master_endo %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

table(subset_data$Cluster)

subset_data$Cluster
```

site 
```{r}
library()
dat_de_master_endo %>%
  group_by(`Cluster`) %>%
  count(`Site Abbreviation`) %>%
  mutate(percent = n/sum(n) * 100)

result <- dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(`Site Abbreviation`) %>%
  mutate(percent = n/sum(n) * 100) %>%
  ungroup()

print(result)

# Calculate the chi-square test
p_value <- chisq.test(result$n, correct = FALSE)$p.value

print(p_value)
```





CV risk
```{r}
dat_de_master_endo %>%
  group_by(`Cluster`) %>%
  count(CV_Risk) %>%
  mutate(percent = n/sum(n) * 100)

## Pvalue
result <- dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(CV_Risk) %>%
  mutate(percent = n/sum(n) * 100) %>%
  ungroup()

print(result)

# Calculate the chi-square test
p_value <- chisq.test(result$n, correct = FALSE)$p.value

print(p_value)

# Perform Fisher's exact test for proportions
fisher_result <- fisher.test(table(subset_data$Cluster, subset_data$CV_Risk))


# Extract the p-value from the test result
p_value <- fisher_result$p.value
print(p_value)

```


Sex
```{r}
## for 3 endotypes 
dat_de_master_endo %>%
  group_by(`Cluster`) %>%
  count(dem_sex) %>%
  mutate(percent = n/sum(n) * 100)

result <- dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(dem_sex) %>%
  mutate(percent = n/sum(n) * 100) %>%
  ungroup()

library(dplyr)


# Create a contingency table for chi-square test
cont_table <- table(result$Cluster, result$dem_sex)
print(cont_table)

# Perform chi-square test
chi_sq_result <- chisq.test(cont_table)

# Extract the p-value
p_value <- chi_sq_result$p.value

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype B"
dat_de_master_endo$Cluster <-  as.character(dat_de_master_endo$Cluster)
str(subset_data$Cluster)

library(dplyr)
subset_data <- dat_de_master_endo %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$dem_sex)
print(cont_table)

# Perform Chisq's exact test ############   ,,,,n=5 use chisq
chisq_test_result <- chisq.test(cont_table)

# Extract the p-value
print(chisq_test_result)

```


Age
```{r}
## for all the 3 endotypes
dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(dem_age),
            q1 = quantile(dem_age, probs = 0.25),
            q3 = quantile(dem_age, probs = 0.75),
            iqr = IQR(dem_age))
####Pvalue
library(dplyr)
## Age is normally distributed; since the sample size might be small we performed a shapiro test. 
shapiro.test(dat_de_master_endo$dem_age)

result <- dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(dem_age),
            q1 = quantile(dem_age, probs = 0.25),
            q3 = quantile(dem_age, probs = 0.75),
            iqr = IQR(dem_age))

# Perform ANOVA
anova_result <- aov(dem_age ~ Cluster, data = dat_de_master_endo)

# Extract the p-value from the ANOVA result
p_value <- summary(anova_result)[[1]]$`Pr(>F)`[1]
print(p_value)

## between endotype A and Endotype C
# Perform a two-sample t-test since sample size is not too small i.e >40
t_test_result <- t.test(dem_age ~ Cluster, data = subset_data)

# Extract the p-value
p_value <- t_test_result$p.value
p_value

```

DM duration
```{r}
library(dplyr)

dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(dm_dur_in_years),
            q1 = quantile(dm_dur_in_years, probs = 0.25),
            q3 = quantile(dm_dur_in_years, probs = 0.75),
            iqr = IQR(dm_dur_in_years))

shapiro.test(dat_de_master_endo$dm_dur_in_years) ##normally distributed

# Perform ANOVA
anova_result <- aov(dm_dur_in_years ~ Cluster, data = dat_de_master_endo)

# Extract the p-value from the ANOVA result
p_value <- summary(anova_result)[[1]]$`Pr(>F)`[1]
print(p_value)

## between endotype A and Endotype C
# Perform a two-sample t-test since sample size is not too small i.e >40
wilcoxon_test_result <- wilcox.test(dm_dur_in_years  ~ Cluster, data = subset_data)

# Extract the p-value
p_value <- wilcoxon_test_result$p.value

```


DM complications
Foot amp
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_foot_amp) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_foot_amp))$p.value
p_value

```
Retinopathy 
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_diab_eye) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_diab_eye))$p.value
p_value

```
Kidney disease
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_kid_disease) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_kid_disease))$p.value
p_value
```
Peripheral neuropathy 
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_per_neur) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_per_neur))$p.value
p_value
```
Hypertension
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_hype) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_hype))$p.value
p_value
```
Stroke
```{r}
dat_de_master_endo%>%
  group_by(Cluster)%>%
  count(dm_stroke) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$dm_stroke))$p.value
p_value
```
Medications use
Insulin
```{r}
dat_de_master_endo%>%
   group_by(Cluster)%>%
  count(con_Insulin) %>%
 mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$con_Insulin))$p.value
p_value

#### BY ENDOTYPES A VS C
# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$con_Insulin)

# Perform Fisher's exact test
fisher_test_result <- fisher.test(cont_table)

# Extract the p-value
p_value <- fisher_test_result$p.value

print(p_value)
```

Metformin
```{r}
dat_de_master_endo%>%
   group_by(Cluster)%>%
  count(con_Metformin) %>%
 mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$con_Metformin))$p.value
p_value

#### BY ENDOTYPES A VS C
# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$con_Metformin)

# Perform Fisher's exact test
fisher_test_result <- fisher.test(cont_table)

# Extract the p-value
p_value <- fisher_test_result$p.value

print(p_value)

```
SUs
```{r}
dat_de_master_endo%>%
   group_by(Cluster)%>%
  count(con_su_der) %>%
 mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$con_su_der))$p.value
p_value

# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$con_su_der)

# Perform Fisher's exact test
fisher_test_result <- fisher.test(cont_table)

# Extract the p-value
p_value <- fisher_test_result$p.value

print(p_value)
```

Other dm medication
```{r}
dat_de_master_endo%>%
   group_by(Cluster)%>%
  count(con_other_aDmed) %>%
 mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$con_other_aDmed))$p.value
p_value

# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$con_other_aDmed)

# Perform Fisher's exact test
fisher_test_result <- fisher.test(cont_table)

# Extract the p-value
p_value <- fisher_test_result$p.value

print(p_value)
```
No use of any medications
```{r}
dat_de_master_endo%>%
   group_by(Cluster)%>%
  summarise(count_no_medication = sum(con_Metformin == "No" & con_Insulin == "No" & con_su_der == "No" & con_other_aDmed == "No"),
            percent = count_no_medication / n() * 100)

.............................................................................

result <- dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(count_no_medication = sum(con_Metformin == "No" & con_Insulin == "No" & con_su_der == "No" & con_other_aDmed == "No"),
            percent = count_no_medication / n() * 100)

p_value <- chisq.test(result$count_no_medication)$p.value

p_value


```


HbA1c and HbA1c categorical
```{r}
dat_de_master_endo$HbA1c_value <- as.numeric(dat_de_master_endo$HbA1c_value)

sum(is.na(dat_de_master_endo$HbA1c_value)) #122 have missing values they are cohort
dat_de_master_endo$HbA1c_value[dat_de_master_endo$HbA1c_value == -99] <- NA -99 #-99 is when the test was not done

sum(is.na(dat_de_master_endo$HbA1c_value)) #138 total missing

##summary statistics for Hb A1c

mean_hba1c <- mean(dat_de_master_endo$HbA1c_value, na.rm = TRUE)
mean_hba1c
quantile(dat_de_master_endo$HbA1c_value, probs = c(0.25, 0.75), na.rm = TRUE)
IQR_hba1c <- IQR(dat_de_master_endo$HbA1c_value, na.rm = TRUE)
IQR_hba1c

library(dplyr)

dat_de_master_endoHbA1c <- dat_de_master_endo %>% 
  filter(!is.na(HbA1c_value)) %>% # drop missing values
  mutate(HbA1c_cat = case_when(
    HbA1c_value < 7 ~ "<7",
    HbA1c_value >= 7 & HbA1c_value <= 10 ~ "7-10",
    HbA1c_value > 10 ~ ">10"
  )) # create new variable with 3 categories 103 patients

dat_de_master_endoHbA1c %>%
  group_by(Cluster) %>%
  count(HbA1c_cat) %>%
  mutate(percent = n/sum(n) * 100)

result <- dat_de_master_endoHbA1c %>%
  group_by(Cluster) %>%
  count(HbA1c_cat) %>%
  mutate(percent = n/sum(n) * 100)

result <- subset(result, Cluster %in% c("Endotype A", "Endotype C"))

# Perform chi-square test
ftest <- fisher.test(table(result$Cluster, result$HbA1c_cat))

# Extract the p-value
p_value <- ftest$p.value

p_value

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype C"
subset_data2 <- dat_de_master_endoHbA1c %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

# Create a contingency table
cont_table <- table(subset_data2$HbA1c_cat, subset_data2$Cluster)

# Perform Fisher's exact test for proportions
fisher_result <- fisher.test(cont_table)

# Extract the p-value from the test result
p_value <- fisher_result$p.value

print(p_value)

...........................................................................................
dat_de_master_endoHbA1c %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(HbA1c_value),
            q1 = quantile(HbA1c_value, probs = 0.25),
            q3 = quantile(HbA1c_value, probs = 0.75),
            iqr = IQR(HbA1c_value))

# Check for normality
shapiro.test(dat_de_master_endoHbA1c$HbA1c_value) ##nornally distributed

model <- lm(HbA1c_value ~ Cluster, data = dat_de_master_endoHbA1c)
anova(model)

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype C"
subset_data2 <- dat_de_master_endoHbA1c %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

# Perform a Wilcoxon rank-sum test
wilcoxon_test_result <- wilcox.test(HbA1c_value ~ Cluster, data = subset_data2, exact = FALSE)### since there are ties I add the exact=FALSE


# Extract the p-value
p_value <- wilcoxon_test_result$p.value

# Print the p-value
print(p_value)

```


Visceral fat and total fat
```{r}

library(dplyr)
sum(is.na(dat_de_master_endo$visceralfat_lev)) 
dat_de_master_endo$visceralfat_lev[dat_de_master_endo$visceralfat_lev%in% c(-95, -96, -99)] <- NA
sum(is.na(dat_de_master_endo$visceralfat_lev)) ##5 variables with -95 or -99 or -96

dat_de_master_endo$visceralfat_lev <- as.numeric(dat_de_master_endo$visceralfat_lev)

dat_de_master_endo %>%
  filter(!is.na(visceralfat_lev)) %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(visceralfat_lev),
            q1 = quantile(visceralfat_lev, probs = 0.25),
            q3 = quantile(visceralfat_lev, probs = 0.75),
            iqr = IQR(visceralfat_lev))

# Check for normality
shapiro.test(dat_de_master_endo$visceralfat_lev)

model <- lm(visceralfat_lev ~ Cluster, data = dat_de_master_endo)
anova(model)

# Perform Kruskal-Wallis test (due to similarity between distribution of the two groups but abnormal sitributon)
kruskal_test_result <- kruskal.test(visceralfat_lev~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)

```

Total body fat
```{r}
sum(is.na(dat_de_master_endo $body_fat)) 
dat_de_master_endo $body_fat[dat_de_master_endo $body_fat%in% c(-95, -96, -99)] <- NA
sum(is.na(dat_de_master_endo $body_fat)) ##5 variables with -95 or -99 or -96

dat_de_master_endo $body_fat <- as.numeric(dat_de_master_endo $body_fat)

dat_de_master_endo %>%
  filter(!is.na(body_fat)) %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(body_fat),
            q1 = quantile(body_fat, probs = 0.25),
            q3 = quantile(body_fat, probs = 0.75),
            iqr = IQR(body_fat))

# Check for normality
shapiro.test(dat_de_master_endo$body_fat)

model <- lm(body_fat ~ Cluster, data = dat_de_master_endo)
anova(model)

# Perform Kruskal-Wallis test (due to similarity between distribution of the two groups but abnormal sitributon)
kruskal_test_result <- kruskal.test(body_fat~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)


```


Anthropometrics
```{r}
### MUAC

dat_de_master_endo$anth_upperarm <- as.numeric(dat_de_master_endo$anth_upperarm)

dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(anth_upperarm),
            q1 = quantile(anth_upperarm, probs = 0.25),
            q3 = quantile(anth_upperarm, probs = 0.75),
            iqr = IQR(anth_upperarm))

# Check for normality
shapiro.test(dat_de_master_endo$anth_upperarm) ## not normally distributed

# Perform ANOVA for age
model <- lm(anth_upperarm ~ Cluster, data = dat_de_master_endo)
anova(model)

### ENdotype A vs C 

# Perform Kruskal-Wallis test (due to similarity between distribution of the two groups but abnormal sitributon)
kruskal_test_result <- kruskal.test(anth_upperarm~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)

```

WHR 
```{r}

dat_de_master_endo$anth_Waisthipratio<- as.numeric(dat_de_master_endo$anth_Waisthipratio)

dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(anth_Waisthipratio),
            q1 = quantile(anth_Waisthipratio, probs = 0.25),
            q3 = quantile(anth_Waisthipratio, probs = 0.75),
            iqr = IQR(anth_Waisthipratio))
# Check for normality

shapiro.test(dat_de_master_endo$anth_Waisthipratio) # normally distributed

# Perform ANOVA 
model <- lm(anth_Waisthipratio ~ Cluster, data = dat_de_master_endo)
anova(model)

### ENdotype A vs C 

# Perform Kruskal-Wallis test (due to similarity between distribution)
kruskal_test_result <- kruskal.test(anth_Waisthipratio~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)
```

WHC
```{r}

dat_de_master_endo$anth_waist<- as.numeric(dat_de_master_endo$anth_waist)

dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(anth_waist),
            q1 = quantile(anth_waist, probs = 0.25),
            q3 = quantile(anth_waist, probs = 0.75),
            iqr = IQR(anth_waist))
# Check for normality
shapiro.test(dat_de_master_endo$anth_waist)

# Perform ANOVA 
model <- lm(anth_waist~ Cluster, data = dat_de_master_endo)
anova(model)

### ENdotype A vs C 

# Perform Kruskal-Wallis test (due to similarity between distribution)
kruskal_test_result <- kruskal.test(anth_waist~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)
```

Smoking
```{r}
dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(Smoking) %>%
  mutate(percent = n/sum(n) * 100)
```

Alcohol use
```{r}
dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(Alc_currconsume) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(dat_de_master_endo$Cluster, dat_de_master_endo$Alc_currconsume)$p.value
print(p_value)

### Endotype A vs C
# Create a contingency table
cont_table <- table(subset_data$Cluster, subset_data$Alc_currconsume)

# Perform Fisher's exact test
fisher_test_result <- fisher.test(cont_table)

# Extract the p-value
p_value <- fisher_test_result$p.value

print(p_value)

```

BMI
```{r}
dat_de_master_endo %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(anth_bmi),
            q1 = quantile(anth_bmi, probs = 0.25),
            q3 = quantile(anth_bmi, probs = 0.75),
            iqr = IQR(anth_bmi))

# Check for normality
shapiro.test(dat_de_master_endo$anth_bmi)

# Perform ANOVA 
model <- lm(anth_bmi~ Cluster, data = dat_de_master_endo)
anova(model)

## Between endotype A and C
# Perform Kruskal-Wallis test (due to similarity between distribution of the two groups but abnormal sitributon)
kruskal_test_result <- kruskal.test(anth_bmi~ Cluster, data = subset_data)

# Extract the p-value
p_value <- kruskal_test_result$p.value

print(p_value)


#### BMI cat proportions
dat_de_master_endo%>%
  group_by(Cluster) %>%
  count(BMI_cat) %>%
  mutate(percent = n/sum(n) * 100)


# Perform Fisher's exact test for proportions
fisher_result <- fisher.test(table(subset_data$Cluster, subset_data$BMI_cat))

# Extract the p-value from the test result
p_value <- fisher_result$p.value
print(p_value)

```


##Additional biochemistry tests

Cpeptide
```{r}
### I have some missing values I know so I want to ignore those
unique_Cpeptide <- unique(dat_de_master_endo$Cpeptide)
print(unique_Cpeptide)

###remove insulin users 
insulin_cpeptide <- dat_de_master_endo %>%
  filter(Cpeptide %in% unique_Cpeptide, con_Insulin == "No") ###205

# Remove na
insulin_cpeptide1 <- insulin_cpeptide %>%
  filter(!is.na(Cpeptide))#200 IDs

insulin_cpeptide1%>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(Cpeptide, na.rm = TRUE),
            q1 = quantile(Cpeptide, probs = 0.25, na.rm = TRUE),
            q3 = quantile(Cpeptide, probs = 0.75, na.rm = TRUE),
            iqr = IQR(Cpeptide, na.rm = TRUE))

model <- lm(Cpeptide ~ Cluster, data = insulin_cpeptide1)
anova(model)


insulin_cpeptide%>%
  group_by(dem_sex) %>%
  summarise(n = n(),
            median = median(Cpeptide, na.rm = TRUE),
            q1 = quantile(Cpeptide, probs = 0.25, na.rm = TRUE),
            q3 = quantile(Cpeptide, probs = 0.75, na.rm = TRUE),
            iqr = IQR(Cpeptide, na.rm = TRUE))

insulin_cpeptide1 %>%
  group_by(Cluster) %>%
  summarise(count_lt_0.2 = sum(Cpeptide < 0.2, na.rm = TRUE),
            percentage_lt_0.2 = sum(Cpeptide < 0.2, na.rm = TRUE) / n() * 100)

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype B"
subset_data4 <- insulin_cpeptide1 %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

subset_data4$Cluster

# Perform Fisher's exact test for proportions
wilcox_result <- wilcox.test(table(subset_data4$Cluster, subset_data4$Cpeptide))

# Extract the p-value from the test result
p_value <- wilcox_result$p.value
print(p_value)


```


Creatinine
```{r}
##there are 0 variables in the dataset however; its NA
dat_de_master_endo$Kreatinine1

dat_de_master_endokre <- dat_de_master_endo %>%
  filter(!is.na(Kreatinine1))#238

dat_de_master_endokre$Kreatinine1

dat_de_master_endokre %>%
  mutate(Kreatinine1 = ifelse(Kreatinine1 == 0.00, NA, Kreatinine1)) %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(Kreatinine1, na.rm = TRUE),
            q1 = quantile(Kreatinine1, probs = 0.25, na.rm = TRUE),
            q3 = quantile(Kreatinine1, probs = 0.75, na.rm = TRUE),
            iqr = IQR(Kreatinine1, na.rm = TRUE))

shapiro.test(dat_de_master_endokre$Kreatinine1) #normally distributed

model <- lm(Kreatinine1 ~ Cluster, data = dat_de_master_endokre)
anova(model)

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype B"
subset_data5 <- dat_de_master_endokre %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

# Perform Fisher's exact test for proportions
wilcox_result <- wilcox.test(table(subset_data5$Cluster, subset_data5$Kreatinine1))

# Extract the p-value from the test result
p_value <- wilcox_result$p.value
print(p_value)
```

Uraat
```{r}
library(dplyr)

dat_de_master_endo$Uraat

dat_de_master_endoUra <- dat_de_master_endo %>%
  filter(!is.na(Uraat))#233

dat_de_master_endoUra %>%
  group_by(Cluster) %>%
  summarise(n = n(),
            median = median(Uraat, na.rm = TRUE),
            q1 = quantile(Uraat, probs = 0.25, na.rm = TRUE),
            q3 = quantile(Uraat, probs = 0.75, na.rm = TRUE),
            iqr = IQR(Uraat, na.rm = TRUE))

shapiro.test(dat_de_master_endoUra$Uraat)# normally distributed

model <- lm(Uraat ~ Cluster, data = dat_de_master_endoUra)
anova(model)

## between endotype A and Endotype C
# Subset the data for "Endotype A" and "Endotype B"
subset_data6 <- dat_de_master_endoUra %>%
  filter(Cluster %in% c("Endotype A", "Endotype C"))

# Perform Fisher's exact test for proportions
t_result <- t.test(table(subset_data6$Cluster, subset_data6$Kreatinine1))

# Extract the p-value from the test result
p_value <- t_result$p.value
print(p_value)

```

### #### distribution of cpeptide and uric acid
```{r}
library(ggplot2)

options(repr.plot.width = 8, repr.plot.height = 5)

# Filter missing values
filtered_data2 <- insulin_cpeptide[complete.cases(insulin_cpeptide$Uraat, insulin_cpeptide$Cpeptide), ]
str(filtered_data2)

filtered_data2$Cpeptide <- as.numeric(filtered_data2$Cpeptide)

ggplot(filtered_data2, aes(x = Uraat)) +
  geom_density(alpha = 0.5, color = "seagreen") +
  geom_density(aes(x = Cpeptide), alpha = 0.5, color = "purple") +
  geom_vline(xintercept = c(0.46, 0.36), linetype = "dotted", color = "black", size = 0.2) +
  labs(title = "Density Plot of Uric Acid and C-peptide Levels",
       x = "Levels",
       y = "Density") +
  theme_classic()
```


#### distribution of creatinine levels by sex
```{r}
library(ggplot2)

options(repr.plot.width = 8, repr.plot.height = 5)

# Filter missing values
filtered_data3 <- Metadata_clean1[complete.cases(Metadata_clean1$Kreatinine1), ]

ggplot(filtered_data3, aes(x = Kreatinine1, fill = dem_sex, color = dem_sex)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(96, 81), linetype = "dotted", color = "black", size = 0.2) +
  labs(title = "Density Plot of Creatinine Levels by Sex",
       x = "Creatinine Levels",
       y = "Density") +
  theme_classic()
```


CKD stage
```{r}

dat_de_master_endo$CKD_Stage
na_count <- sum(is.na(dat_de_master_endo$CKD_Stage))
print(na_count)

dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(CKD_Stage, na.rm = TRUE) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endokre$Cluster, dat_de_master_endokre$CKD_Stage))$p.value
p_value

dat_de_master_endokre$Kreatinine1

# Perform Fisher's exact test for proportions
fisher_result <- fisher.test(table(subset_data$Cluster, subset_data$CKD_Stage))

# Extract the p-value from the test result
p_value <- fisher_result$p.value
print(p_value)
```


LTBI status
```{r}
dat_de_master_endo %>%
  group_by(Cluster) %>%
  count(igra_ltbiconf, na.rm = TRUE) %>%
  mutate(percent = n/sum(n) * 100)

p_value <- chisq.test(table(dat_de_master_endo$Cluster, dat_de_master_endo$igra_ltbiconf))$p.value
p_value

# Perform Fisher's exact test for proportions
fisher_result <- fisher.test(table(subset_data$Cluster, subset_data$igra_ltbiconf))

# Extract the p-value from the test result
p_value <- fisher_result$p.value
print(p_value)
```


