---
title: "Merged data proteomics and metadata"
author: "Lucy"
date: "2023-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Load the merged rds file
```{r}
dat_de_master <- readRDS("Merged.rds")

Olink <- readRDS("PROTID_Olink2.rds")

#library(openxlsx)
#write.xlsx(Olink, file = "D:/Trainings/PhD/Analysis/Olink.xlsx")

colnames(dat_de_master)
```


PCA on all proteins first 
```{r}
library(ggplot2)
library(ggbiplot)
prin_comp <- prcomp(dat_de_master[,c(62:138)], center = T, scale. = T)

  ggbiplot(prin_comp, obs.scale = 1, var.scale = 1, 
                ellipse = F, circle = F, var.axes = F)+
        theme(legend.direction = 'horizontal', 
              legend.position = 'top')+
        theme_bw(base_size=10)
```


Heat map 
```{r}
### Unsupervised clustering analysis
install.packages("ComplexHeatmap")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)
#BiocManager::install("circlize", force = TRUE)
library(circlize)

mat <- as.matrix(t(scale(Olink[,-c(1)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))

heatmap <- Heatmap(
  mat, 
  name = "NPX", 
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.5),
  width = unit(90, "mm"),  # Increase the width value
  height = unit(90, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 6)
)

heatmap <- draw(
  heatmap, 
  heatmap_legend_side = "right", 
  annotation_legend_side = "left"
)
```


4 endotypes 
```{r}
library(circlize)
library(ComplexHeatmap)

mat <- as.matrix(t(scale(Olink[,-c(1)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))

# Perform hierarchical clustering
hc_rows <- hclust(dist(mat))
hc_cols <- hclust(dist(t(mat)))

# Identify the clusters
row_clusters <- cutree(hc_rows, k = 3)
col_clusters <- cutree(hc_cols, k = 3)

# Create a matrix for row and column annotations
row_anno <- matrix(row_clusters, nrow = nrow(mat))
col_anno <- matrix(col_clusters, nrow = ncol(mat))

# Identify the columns belonging to Cluster 2
cluster2_cols <- which(col_anno == 2)



# Create a custom column annotation function
column_annotation_fun <- function(density) {
  anno <- HeatmapAnnotation(
    df = data.frame(density = density),
    show_legend = FALSE,
    width = unit(1, "cm"),
    height = unit(5, "cm"),
    col = col_fun
  )
  anno
}

# Create the heatmap with custom column annotation
heatmap <- Heatmap(
  mat,
  name = "NPX",
  col = col_fun,
  rect_gp = gpar(col = "grey", lwd = 0.5),
  width = unit(90, "mm"),
  height = unit(90, "mm"),
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 6),
  top_annotation = column_annotation_fun(col_anno)
)

heatmap <- draw(
  heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "left"
)

# Display the updated heatmap
draw(heatmap)
```





Dividing the dataset into the three endotypes 
```{r}
library(ComplexHeatmap)

mat_transposed <- t(mat)


# Perform hierarchical clustering on your data
hclust_result <- hclust(dist(mat_transposed))

# Determine cluster labels
k <- 3  # Number of clusters (adjust as needed)
cluster_labels <- cutree(hclust_result, k)


# Add cluster labels as a column in the dataset
Olink$Cluster <- cluster_labels

colnames(Olink)
print(Olink$Cluster)

###Renaming clusters based on the heat map
levels(Olink$Cluster) <- c("Endotype B", "Endotype C", "Endotype A")
Olink$Cluster <- factor(Olink$Cluster)

print(Olink$Cluster)

```

Now I replot the heat map and see wether it is truelly reflected
```{r}
### Unsupervised clustering analysis
library(ComplexHeatmap)
#BiocManager::install("circlize", force = TRUE)
library(circlize)

mat2 <- as.matrix(t(scale(Olink[,-c(1,79)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))

ha <- HeatmapAnnotation(
  Cluster = as.factor(Olink$Cluster),
  col = list(Cluster = c("Endotype A" = "darkgreen", "Endotype B" = "orange", "Endotype C" = "darkred")))

heatmap <- Heatmap(
  mat2, 
  name = "NPX",
  top_annotation = ha,
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.02),
  width = unit(90, "mm"),  # Increase the width value
  height = unit(90, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 4) 
)

heatmap <- draw(heatmap, 
                  heatmap_legend_side = "right", 
                  annotation_legend_side = "left")
    
```



```{r}
#### with site as a covariate

ha <- HeatmapAnnotation(
  Site = dat_de_master$`Site Abbreviation`,
  col = list(
    Site= c(MBE = "lightblue", MOS = "seagreen")
  )
)

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))

heatmap <- Heatmap(
  mat, 
  name = "NPX",
  top_annotation = ha,
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.5),
  width = unit(90, "mm"),  # Increase the width value
  height = unit(90, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 6)
)

heatmap <- draw(
  heatmap, 
  heatmap_legend_side = "right", 
  annotation_legend_side = "left"
)
```
Filter one site only to see wether the endotypes exist: Moshi
```{r}
dat_de_master_endo <- readRDS("dat_de_master_endo.rds")

# Filter out rows where 'Site Abbreviation' is 'MBE'
dat_de_filtered <- subset(dat_de_master_endo, `Site Abbreviation` != "MBE")

library(circlize)

library(ComplexHeatmap)

colnames(dat_de_filtered)

mat <- as.matrix(t(scale(dat_de_filtered[,-c(1:62,140)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))


ha <- HeatmapAnnotation(
  Age = dat_de_filtered$dem_age,
  Sex = dat_de_filtered$dem_sex,
  BMI = dat_de_filtered$anth_bmi,
  DM_duration = dat_de_filtered$dm_dur_in_years,
  LTBI_status = dat_de_filtered$igra_ltbiconf,
  CV_risk = dat_de_filtered$WHO_ISH_Risk,
  Site= dat_de_filtered$`Site Abbreviation`,
  
  col = list(
    Age = colorRamp2(c(min(dat_de_filtered$dem_age), max(dat_de_filtered$dem_age)), c("white", "#0571b0")),
    Sex = c(Male = "#1F78B4", Female = "#FB9A99"),
    BMI = colorRamp2(c(min(dat_de_filtered$anth_bmi), max(dat_de_filtered$anth_bmi)), c("white", "red")),
    DM_duration = colorRamp2(c(min(dat_de_filtered$dm_dur_in_years), max(dat_de_filtered$dm_dur_in_years)), c("white", "darkred")),
    LTBI_status = c(Yes = "darkred", No = "seagreen"), 
    CV_risk = c("<10%" = "#edf8e9", "10% to <20%" = "#fecc5c", "20% to <30%" = "#fd8d3c", "30% to <40%" = "#cc4c02", ">=40%" = "#e31a1c"), Site= c(MBE = "lightblue", MOS = "seagreen")
  )
)




heatmap <- Heatmap(
  mat, 
  name = "NPX",
  top_annotation = ha,
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.02),
  width = unit(100, "mm"),  # Increase the width value
  height = unit(100, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 4) 
)

heatmap <- draw(heatmap, 
                  heatmap_legend_side = "right", 
                  annotation_legend_side = "left")

```

Filter one site only to see wether the endotypes exist: Mbeya
```{r}
dat_de_master_endo <- readRDS("dat_de_master_endo.rds")

# Filter out rows where 'Site Abbreviation' is 'M0S'
dat_de_filtered_mbe <- subset(dat_de_master_endo, `Site Abbreviation` != "MOS")

library(circlize)

library(ComplexHeatmap)

colnames(dat_de_filtered_mbe)

mat <- as.matrix(t(scale(dat_de_filtered_mbe[,-c(1:62,140)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))


ha <- HeatmapAnnotation(
  Age = dat_de_filtered_mbe$dem_age,
  Sex = dat_de_filtered_mbe$dem_sex,
  BMI = dat_de_filtered_mbe$anth_bmi,
  DM_duration = dat_de_filtered_mbe$dm_dur_in_years,
  LTBI_status = dat_de_filtered_mbe$igra_ltbiconf,
  CV_risk = dat_de_filtered_mbe$WHO_ISH_Risk,
  Site= dat_de_filtered_mbe$`Site Abbreviation`,
  col = list(
    Age = colorRamp2(c(min(dat_de_filtered_mbe$dem_age), max(dat_de_filtered_mbe$dem_age)), c("white", "#0571b0")),
    Sex = c(Male = "#1F78B4", Female = "#FB9A99"),
    BMI = colorRamp2(c(min(dat_de_filtered_mbe$anth_bmi), max(dat_de_filtered_mbe$anth_bmi)), c("white", "red")),
    DM_duration = colorRamp2(c(min(dat_de_filtered_mbe$dm_dur_in_years), max(dat_de_filtered_mbe$dm_dur_in_years)), c("white", "darkred")),
    LTBI_status = c(Yes = "darkred", No = "seagreen"), 
    CV_risk = c("<10%" = "#edf8e9", "10% to <20%" = "#fecc5c", "20% to <30%" = "#fd8d3c", "30% to <40%" = "#cc4c02", ">=40%" = "#e31a1c"), Site= c(MBE = "lightblue", MOS = "seagreen")
  )
)



heatmap <- Heatmap(
  mat, 
  name = "NPX",
  top_annotation = ha,
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.02),
  width = unit(100, "mm"),  # Increase the width value
  height = unit(100, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 4) 
)

heatmap <- draw(heatmap, 
                  heatmap_legend_side = "right", 
                  annotation_legend_side = "left")

# Specify the directory for saving the figure
dir_figures <- "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Nightingale_downstream_analysis/Figures"

# Use ggsave to save the plot
library(ggplot2)
ggsave(
  filename = file.path(dir_figures, "Figure4.pdf"),
  plot = heatmap,
  height = 20,
  width = 22
)

```

Save the rds file 
```{r}
saveRDS(Olink, "Olink_endotypes.rds")
```


Merge the endotypes and the Metadata
```{r}
###Merge proteomic data with metadata (total 241 participants)
   Metadata_clean241 <- readRDS("D:/Trainings/PhD/Analysis/Metadata_clean241.rds")


  dat_de_master_endo <- merge(Metadata_clean241,Olink,by.0="PID") 
  saveRDS(dat_de_master_endo, "dat_de_master_endo.rds")
  
  colnames(dat_de_master_endo)

```


Heat map with the DM specific variables that had associations
```{r}
library(circlize)

library(ComplexHeatmap)

colnames(dat_de_master_endo)

mat <- as.matrix(t(scale(dat_de_master_endo[,-c(1:62,140)])))

col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue4", "white", "red4"))


ha <- HeatmapAnnotation(
  Age = dat_de_master_endo$dem_age,
  Sex = dat_de_master_endo$dem_sex,
  BMI = dat_de_master_endo$anth_bmi,
  DM_duration = dat_de_master_endo$dm_dur_in_years,
  LTBI_status = dat_de_master_endo$igra_ltbiconf,
  CV_risk = dat_de_master_endo$WHO_ISH_Risk,
  col = list(
    Age = colorRamp2(c(min(dat_de_master_endo$dem_age), max(dat_de_master_endo$dem_age)), c("white", "#0571b0")),
    Sex = c(Male = "#1F78B4", Female = "#FB9A99"),
    BMI = colorRamp2(c(min(dat_de_master_endo$anth_bmi), max(dat_de_master_endo$anth_bmi)), c("white", "red")),
    DM_duration = colorRamp2(c(min(dat_de_master_endo$dm_dur_in_years), max(dat_de_master_endo$dm_dur_in_years)), c("white", "darkred")),
    LTBI_status = c(Yes = "darkred", No = "seagreen"), 
    CV_risk = c("<10%" = "#edf8e9", "10% to <20%" = "#fecc5c", "20% to <30%" = "#fd8d3c", "30% to <40%" = "#cc4c02", ">=40%" = "#e31a1c")
  )
)




heatmap <- Heatmap(
  mat, 
  name = "NPX",
  top_annotation = ha,
  col = col_fun,
  column_split = 3,
  rect_gp = gpar(col = "grey", lwd = 0.02),
  width = unit(100, "mm"),  # Increase the width value
  height = unit(100, "mm"),  # Increase the height value
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 4) 
)

heatmap <- draw(heatmap, 
                  heatmap_legend_side = "right", 
                  annotation_legend_side = "left")
```

DEA analysis
```{r}
#DEA between the endotypes
library(limma)
dim(dat_de_master_endo)
```

 #Generating the design matrix 
```{r}
    # Design matrix without covariate
    dat_de_pr <- dat_de_master_endo$Cluster
    Cluster <- dat_de_pr
    
    design_cov <- model.matrix(~Cluster) 
    colnames(design_cov) <- c("Intercept", "ClusterEndotypeA", "ClusterEndotypeB")
    dim(design_cov)
```

 # Generate the protein expression matrix 
```{r}
colnames(dat_de_master_endo)
    # Transform NPX data for proteins to represent rows - no group names
    # Remove specific columns so you only have the protein measurements
    dat_de <- dat_de_master_endo[,-c(1:61,139)]
    dim(dat_de)
```
```{r}
dat_de_tr <- as.matrix(t(dat_de))  # Transpose data

    #Fitting models for differentially expressed proteins
    # LMFIT ASSUMES YOUR VALUES ARE ALREADY LOG TRANSFORMED 
    fit1_cov <- lmFit(dat_de_tr, design_cov)
    cont.matrix_cov <- makeContrasts("ClusterEndotypeA", levels = design_cov)
    fit2_cov <- contrasts.fit(fit1_cov,cont.matrix_cov)
    fit2_cov <- eBayes(fit2_cov)
```


```{r}
 # Display "N" number of differentially expressed proteins
    # Use this object for volcano plot 
    DEP_cov_2 <- topTable(fit2_cov, coef = "ClusterEndotypeA", 
                          adjust = "BH", number = ncol(dat_de)) 
    # number = total number of proteins
    
    #Display number of proteins significantly expressed between sex
    results_cov <- decideTests(fit2_cov)
    
    # Display the number of Up and Down 
    summary(results_cov)
```

