---
title: "Script for data analysis"
author: "Lucy Mrema"
date: "2026-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
dat_de_master<- readRDS("C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Data/Proteomic_Analysis/dat_de_master_endo.rds")
colnames(dat_de_master)


Lipidomics_merged_data1 <- readRDS("C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Nightingale_downstream_analysis/Lipidomics_merged_data1_numeric.rds")

colnames(Lipidomics_merged_data1)


```


Inner join/ Merge teh two datasets based on PID
```{r}
merged_data <- merge(dat_de_master, Lipidomics_merged_data1, by = "PID", all = FALSE)

View(merged_data)
head(merged_data)

print(merged_data)
```


Now save the new file with lipids and olink
```{r}
# Save the merged dataset
saveRDS(merged_data, file = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Data/Proteomic_Analysis/dat_de_master_lipid.rds")

```

Re-open it 
```{r fig.width=4, fig.height=5}
dat_de_master_lipid <- readRDS("C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Data/Proteomic_Analysis/dat_de_master_lipid.rds")


library(dplyr)

dat_de_master_lipid %>%
  group_by(dem_sex) %>%
  summarise(
    mean_CFI = mean(CentralFatIndex, na.rm = TRUE),
    median_CFI = median(CentralFatIndex, na.rm = TRUE),
    p25_CFI = quantile(CentralFatIndex, 0.25, na.rm = TRUE),
    p75_CFI = quantile(CentralFatIndex, 0.75, na.rm = TRUE)
  )

library(ggplot2)

# Create the plot
ggplot(dat_de_master_lipid, aes(x = CentralFatIndex, fill = dem_sex)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 30, color = "black") +
  facet_wrap(~dem_sex, scales = "free_y") +
  labs(
    title = "Distribution of Central Fat Index by Sex",
    x = "Central Fat Index",
    y = "Count"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Female" = "#FF69B4", "Male" = "#4682B4"))  # Pink for female, blue for male

# Save it
ggsave("central_fat_index_histogram.png", plot = p, width = 8, height = 5, dpi = 300)


library(ggplot2)

ggplot(dat_de_master_lipid, aes(x = dem_sex, y = CentralFatIndex, fill = dem_sex)) +
  geom_boxplot(outlier.color = "black", outlier.shape = 10, alpha = 0.7, width = 0.4) +
  labs(
    title = "Central Fat Index Distribution by Sex",
    x = "Sex",
    y = "Central Fat Index"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Female" = "#FF69B4", "Male" = "#4682B4")) +
  theme(legend.position = "none")

```

Load required packages 
```{r}
install.packages(c("broom","corpcor","doParallel","ggrepel","glmnet","igraph","magrittr","openxlsx","parallel","pdist","pROC","RColorBrewer","readxl","rsample","tableone","tidyverse"))

```

## Subsetting measurements into categories

```{r}
# -----------------------------
# 1. Define biomarker groups by absolute positions in dat_de_master_lipid
# -----------------------------
Amino_acids <- grep("^(Ala|Gln|Gly|His)$", colnames(dat_de_master_lipid), value = TRUE)
Branched_chain_amino_acids <- grep("^(Ile|Leu|Val)$", colnames(dat_de_master_lipid), value = TRUE)
Aromatic_amino_acids <- grep("^(Phe|Tyr)$", colnames(dat_de_master_lipid), value = TRUE)
Glycolysis_related_metabolites <- grep("^(Glucose.y|Lactate)$", colnames(dat_de_master_lipid), value = TRUE)
Inflammation <- grep("^GlycA$", colnames(dat_de_master_lipid), value = TRUE)
Apolipoproteins <- grep("^(ApoA1|ApoB|ApoB/ApoA1)$", colnames(dat_de_master_lipid), value = TRUE)
Cholesterol <- grep("^(Total-C|VLDL-C|Clinical.LDL-C|LDL-C)$", colnames(dat_de_master_lipid), value = TRUE)
Fatty_acids <- grep("^(Total-FA|Omega-3|Omega-6|PUFA|MUFA|SFA|LA|DHA)$", colnames(dat_de_master_lipid), value = TRUE)
# Remove fatty acid ratios from analysis
# Fatty_acids_ratios <- grep("(%$|PUFA/MUFA|Omega-6/Omega-3)", colnames(dat_de_master_lipid), value = TRUE)
Fluid_balance <- grep("^(Citrate|bOHbutyrate|Acetate|Acetoacetate|Acetone|Creatinine|Albumin)$", colnames(dat_de_master_lipid), value = TRUE)
Triglycerides <- grep("^Total-TG$", colnames(dat_de_master_lipid), value = TRUE)

# -----------------------------
# 2. Combine all selected biomarkers
# -----------------------------
selected_biomarkers <- c(
  Amino_acids,
  Branched_chain_amino_acids,
  Aromatic_amino_acids,
  Glycolysis_related_metabolites,
  Inflammation,
  Apolipoproteins,
  Cholesterol,
  Fatty_acids,
  Fluid_balance,
  Triglycerides
)

# -----------------------------
# 3. Assign group labels to each biomarker
# -----------------------------
assign_group <- function(group_name, biomarkers) {
  setNames(rep(group_name, length(biomarkers)), biomarkers)
}

biomarker_group <- c(
  assign_group("Amino acids", Amino_acids),
  assign_group("Branched-chain amino acids", Branched_chain_amino_acids),
  assign_group("Aromatic amino acids", Aromatic_amino_acids),
  assign_group("Glycolysis-related metabolites", Glycolysis_related_metabolites),
  assign_group("Inflammation", Inflammation),
  assign_group("Apolipoproteins", Apolipoproteins),
  assign_group("Cholesterol", Cholesterol),
  assign_group("Fatty acids", Fatty_acids),
  assign_group("Fluid balance", Fluid_balance),
  assign_group("Triglycerides", Triglycerides)
)

# Remove duplicated biomarker names
biomarker_group <- biomarker_group[!duplicated(names(biomarker_group))]

# -----------------------------
# 4. Create annotation dataframe
# -----------------------------
annotation_col_full <- data.frame(Group = biomarker_group)
rownames(annotation_col_full) <- names(biomarker_group)

# Keep only selected & annotated biomarkers
annotated_biomarkers <- intersect(selected_biomarkers, names(biomarker_group))
annotation_col <- annotation_col_full[annotated_biomarkers, , drop = FALSE]

# Order annotation dataframe by group
annotation_col <- annotation_col[order(annotation_col$Group), , drop = FALSE]
ordered_biomarkers <- rownames(annotation_col)
# -----------------------------
# 5. Subset the data for selected biomarkers
# -----------------------------
dat_subset <- dat_de_master_lipid[, ordered_biomarkers, drop = FALSE]

# -----------------------------
# 6. Transpose so rows = biomarkers, cols = patients
# -----------------------------
dat_scaled <- t(scale((dat_subset)))

# -----------------------------
# 8. Load libraries
# -----------------------------
library(pheatmap)
library(RColorBrewer)

# -----------------------------
# 9. Define annotation for rows (biomarkers)
# -----------------------------
annotation_row <- data.frame(`Biomarker group` = annotation_col$Group)
rownames(annotation_row) <- rownames(annotation_col)

# Define colors for biomarker groups
group_levels <- unique(annotation_row$`Biomarker group`)
group_colors <- setNames(colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(length(group_levels)),
                         group_levels)
ann_colors <- list(`Biomarker group` = group_colors)


# -----------------------------
# 10. Save heatmap as PNG
# -----------------------------
png("heatmap_updated.png", width = 20, height = 10, units = "in", res = 150)

pheatmap(
  dat_scaled,
  annotation_row = annotation_row,   # annotation by biomarker group
  annotation_colors = ann_colors,
  cluster_rows = FALSE,              # keep biomarkers ordered
  cluster_cols = TRUE,               # cluster patients
  show_rownames = TRUE,              # show biomarkers
  show_colnames = TRUE,              # show patients
  fontsize_row = 10,                   # smaller font for biomarkers
  fontsize_col = 10,                   # smaller font for patients
  main = "Z-score Heatmap of Selected Biomarker Groups",
  border_color = NA,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)
)

dev.off()


```
>Heatmap annotated

```{r}
# -----------------------------
# 1. Define biomarker groups
# -----------------------------
biomarker_groups <- list(
  "Amino acids" = c("Ala","Gln","Gly","His"),
  "Branched-chain amino acids" = c("Ile","Leu","Val"),
  "Aromatic amino acids" = c("Phe","Tyr"),
  "Glycolysis-related metabolites" = c("Glucose.y","Lactate"),
  "Inflammation" = c("GlycA"),
  "Apolipoproteins" = c("ApoA1","ApoB","ApoB/ApoA1"),
  "Cholesterol" = c("Total-C","VLDL-C","Clinical.LDL-C","LDL-C"),
  "Fatty acids" = c("Total-FA","Omega-3","Omega-6","PUFA","MUFA","SFA","LA","DHA"),
  "Fluid balance" = c("Citrate","bOHbutyrate","Acetate","Acetoacetate","Acetone","Creatinine","Albumin"),
  "Triglycerides" = c("Total-TG")
)

# -----------------------------
# 2. Map biomarkers to groups
# -----------------------------
biomarker_group <- unlist(
  lapply(names(biomarker_groups), function(g) {
    matches <- grep(paste0("^(", paste(biomarker_groups[[g]], collapse="|"), ")$"),
                    colnames(dat_de_master_lipid), value = TRUE)
    setNames(rep(g, length(matches)), matches)
  })
)

# Remove duplicates if any
biomarker_group <- biomarker_group[!duplicated(names(biomarker_group))]

# -----------------------------
# 3. Order and subset data
# -----------------------------
ordered_biomarkers <- names(sort(biomarker_group))
dat_subset <- dat_de_master_lipid[, ordered_biomarkers, drop = FALSE]

# -----------------------------
# 4. Scale biomarkers (z-scores per biomarker)
# -----------------------------
dat_scaled <- scale(dat_subset) # rows = biomarkers, columns = patients

# -----------------------------
# 5. Define row annotations
# -----------------------------
annotation_row <- data.frame(`Biomarker group` = biomarker_group[ordered_biomarkers])
rownames(annotation_row) <- ordered_biomarkers

# Annotation colors
group_levels <- unique(annotation_col $`Biomarker group`)
group_colors <- setNames(colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(length(group_levels)),
                         group_levels)
ann_colors <- list(`Biomarker group` = group_colors)

# -----------------------------
# 6. Generate heatmap
# -----------------------------
library(pheatmap)
library(RColorBrewer)

png("heatmap_updated.png", width = 20, height = 10, units = "in", res = 150)


pheatmap(
  dat_scaled,
  annotation_col = annotation_col,   # if rows = patients, columns = biomarkers
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = FALSE,  # keep biomarkers ordered
  show_rownames = TRUE,
  show_colnames = TRUE, 
  fontsize_row = 10,                   # smaller font for biomarkers
  fontsize_col = 10,                   # smaller font for patients
  main = "Z-score Heatmap of Selected Biomarker Groups",
  border_color = NA,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)
)

dev.off()
```


Load required packages 
```{r}
library(pheatmap)
library(RColorBrewer)
library(grid)

# Select all lipid biomarkers columns by absolute position
all_biomarkers <- colnames(dat_de_master_lipid)[145:381]

# Extract the data matrix for these biomarkers
lipid_matrix <- dat_de_master_lipid[, all_biomarkers] %>% as.matrix()

# Scale (Z-score) the data by column (biomarker)
lipid_matrix_z <- scale(lipid_matrix)

# Transpose so lipids are rows and patients are columns
lipid_matrix_z <- t(lipid_matrix_z)

# Set column names as patient IDs
if ("PID" %in% colnames(dat_de_master_lipid)) {
  colnames(lipid_matrix_z) <- dat_de_master_lipid$PID
}

# Save heatmap
png("heatmap_lipidscombined.png", width = 16, height = 20, units = "in", res = 150)

# Get 3-class palettes
reds  <- brewer.pal(3, "YlOrRd")   # yellow → orange → red
blues <- brewer.pal(3, "PuBu")     # purple → light blue

# Combine: Blue side for negatives, baby yellow for 0, Red side for positives
my_colors <- c(rev(blues), "#FFFFCC", reds)

pheatmap::pheatmap(
  mat = lipid_matrix_z,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,   # Lipids
  show_colnames = TRUE, # Patients
  cutree_cols = 4,
  color = colorRampPalette(my_colors)(11),
  fontsize_col = 4,
  fontsize_row = 4,
  main = "Z-score Heatmap of All Lipid Biomarkers",
  border_color = NA
)

dev.off()

```


I see three meanful clusters 
```{r}
png("heatmap_lipidscombinedclusters.png", width = 16, height = 18, units = "in", res = 150)

set.seed(123) 

pheatmap::pheatmap(
  mat = lipid_matrix_z,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  cutree_rows = 4,
  color = colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100),
  fontsize_col = 4,
  fontsize_row = 4,
  main = "Z-score Heatmap of All Lipid Biomarkers 3  clusters",
  border_color = NA
)

dev.off()

```
```{r}

# Loading packages.
library(psych)
#library(DESeq2)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
#library(limma)
install.packages("volcano3D")
library(volcano3D)
#library(volcano3Ddata)
library(readxl)
#install.packages("EnvStats")
library(EnvStats)
#install.packages("ggcorrplot")
library(ggcorrplot)
library(knitr)
#install.packages("factoextra")
library(factoextra)
#install.packages("OlinkAnalyze")
library(OlinkAnalyze)
#install.packages("table1")
library(table1)
#install.packages("tidygraph")
library(tidygraph)
#install.packages("igraph")
library(igraph)
#install.packages("ggraph")
library(ggraph)
#install.packages("foreign")
library(foreign)
#install.packages("haven")
library(haven)
#install.packages("iglu")
library(iglu)
library(openxlsx)
library(CorLevelPlot)
#install.packages("flashClust")
library(flashClust)
#install.packages("cowplot")
library(cowplot)
#library(impute)
#library(WGCNA)
#install.packages("reshape2")
library(reshape2)
#install.packages("fs")
library(fs)
library(tidyverse)

```

Load metadata and olink data
```{r} 
dat_de_master <- readRDS("C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Data/Proteomic_Analysis/dat_de_master_lipid.rds")
colnames(dat_de_master)
```

Linear regression for Mbeya and Moshi separately

Moshi
```{r}
library(dplyr)

dat_de_master <- dat_de_master |> 
  rename(Site = "Site Abbreviation")

# Subset data for Mos site
library(dplyr)
dat_mos <- dat_de_master |> filter(Site == "MOS")


# Subset data for Mbe site
dat_mbe <- dat_de_master |> filter(Site == "MBE")


#prepare the data
# Define predictor variables
predictors <- c("anth_bmi", "dem_age","HbA1c_value","Site", "dem_sex","dm_dur_in_years", "visceralfat_lev", "Kreatinine1","CentralFatIndex", "body_fat")

# Extract inflammatory marker column names (columns 63 to 139)
lipid_markers <- names(dat_mos)[145:381]

# Combine all required variables
all_vars <- c(predictors, lipid_markers)

# Check if all variables exist in the dataset
missing <- setdiff(all_vars, names(dat_mos))
if (length(missing) > 0) {
  stop("Missing variables: ", paste(missing, collapse = ", "))
}

# Subset and clean the data
df_sub <- dat_mos [, all_vars]

df_sub$HbA1c_value[df_sub$HbA1c_value == -99] <- NA

# Example data frame
df_sub <- df_sub


# Load broom for tidy model output
library(dplyr)
library(tidyr)


# Convert to long format so we can group by outcome variable
df_long <- df_sub |> 
  pivot_longer(cols = all_of(lipid_markers), 
               names_to = "marker", 
               values_to = "NPX")

# Run regression grouped by each marker (like your `results_lm1`)
results_lm1 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ anth_bmi) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)  |>
    mutate(term = "anth_bmi")
  

results_lm2 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_age) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_age")

results_lm3 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dm_dur_in_years) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dm_dur_in_years")


results_lm4 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ HbA1c_value) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "HbA1c_value")

results_lm5 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ CentralFatIndex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "CentralFatIndex")

results_lm6 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~visceralfat_lev ) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "visceralfat_lev")

results_lm7 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Kreatinine1) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Kreatinine1")

results_lm8 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ body_fat) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "body_fat")

results_lm9 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_sex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_sex")

results_lm <- rbind(results_lm1, results_lm2, results_lm3, results_lm4,results_lm5,results_lm6,results_lm7, results_lm8, results_lm9) |> 
  filter(term != "(Intercept)") |> 
  select(-std.error, -statistic) |> 
  mutate(abs_estimate = abs(estimate),
         direction = estimate / abs_estimate,
         p.adj = p.adjust(p.value, method = "fdr"),
         log10_p_value_and_direction = -log10(p.adj) * direction,
         FDR = cut(log10_p_value_and_direction, 
                   breaks =c(-35, -6, -4, -2.301, -2, -1.301, 1, 1.301, 2, 2.301, 4, 6, 35),
                   labels =c("<0.000001 (neg)", "<0.0001 (neg)", "<0.005 (neg)", "<0.01 (neg)", "<0.05 (neg)", "0", "0",
                             "<0.05 (pos)", "<0.01 (pos)", "<0.005 (pos)", "<0.0001 (pos)", "<0.000001 (pos)")))

# Creating bubble plot.
palette_heatmap <- c("<0.000001 (pos)"="#800026",
                     "<0.0001 (pos)"="#BD0026",
                     "<0.005 (pos)"="#E31A1C",
                     "<0.01 (pos)"="#FC4E2A",
                     "<0.05 (pos)"="#FD8D3C",
                     "non-signif (pos)"="khaki1",
                     "non-signif"="white", 
                     "non-signif (neg)"="lightcyan",
                     "<0.05 (neg)"="#6BAED6",
                     "<0.01 (neg)"="#4292C6",
                     "<0.005 (neg)"="#2171B5",
                     "<0.0001 (neg)"="#08519C",
                     "<0.000001 (neg)"="#08306B")


 results_lm <- results_lm |> 
  mutate(
    marker = factor(marker, levels = rev(unique(marker))),
    term = case_when(
      term == "anth_bmi" ~ "BMI",
      term == "dem_age" ~ "Age",
      term == "HbA1c_value" ~ "HbA1c",
      term == "dm_dur_in_years" ~ "DM duration",
      term == "dem_sex" ~ "Sex (Male)",
      term == "Kreatinine1" ~ "Creatinine",
      term == "visceralfat_lev" ~ "Visceral Fat",
      term == "CentralFatIndex" ~ "Central Fat",
      term == "body_fat" ~ "Total Fat",
      TRUE ~ term
    )
  )
 


# Plot bubble plot
library(ggplot2)

 p <-ggplot( data = results_lm, aes(x = term, 
             y = marker, 
             size = abs_estimate, 
             color = FDR)) + 
  geom_point(shape = 15) + 
  scale_size_continuous(range = c(1, 3.5), breaks = c(0.1, 0.5, 1 , 2, 3, 4, 5, 6,7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17 )) +
  scale_color_manual(values = palette_heatmap, na.value = "grey90") +
  labs(size = "Effect size\n(absolute)") + 
  #facet_grid(~ Phenotype, scale = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 0, size = 10), # changing metadata text size
        axis.text.y = element_text(size = 10),  # changing y axis text size (proteins)
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.3, "line"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(), 
        panel.background = element_rect(fill = "gainsboro"),
        panel.grid = element_blank(),      
        strip.text.y = element_text(size = 5), # changing facet grid text size
        strip.text.x = element_text(size = 10), # changing phenotype text size 
        strip.placement = "outside") + # for placing the strip outside of the axis names
  scale_x_discrete(position = "top")


ggsave("plotMOSlipid.png", plot = p, width = 6, height = 30, units = "in",dpi=600)

```

Mbeya 

```{r}
#prepare the data
# Define predictor variables
predictors <- c("anth_bmi", "dem_age","HbA1c_value","Site", "dem_sex","dm_dur_in_years", "visceralfat_lev", "Kreatinine1","CentralFatIndex", "body_fat")

# Extract inflammatory marker column names (columns 63 to 139)
lipid_markers <- names(dat_mbe)[145:381]

# Combine all required variables
all_vars <- c(predictors, lipid_markers)

# Check if all variables exist in the dataset
missing <- setdiff(all_vars, names(dat_mos))
if (length(missing) > 0) {
  stop("Missing variables: ", paste(missing, collapse = ", "))
}

# Subset and clean the data
df_sub <- dat_mbe [, all_vars]

df_sub$HbA1c_value[df_sub$HbA1c_value == -99] <- NA

# Example data frame
df_sub <- df_sub


# Load broom for tidy model output
library(broom)
library(dplyr)
library(tidyr)


# Convert to long format so we can group by outcome variable
df_long <- df_sub |> 
  pivot_longer(cols = all_of(lipid_markers), 
               names_to = "marker", 
               values_to = "NPX")

# Run regression grouped by each marker (like your `results_lm1`)
results_lm1 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ anth_bmi) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)  |>
    mutate(term = "anth_bmi")
  

results_lm2 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_age) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_age")

results_lm3 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dm_dur_in_years) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dm_dur_in_years")


results_lm4 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ HbA1c_value) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "HbA1c_value")

results_lm5 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ CentralFatIndex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "CentralFatIndex")

results_lm6 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~visceralfat_lev ) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "visceralfat_lev")

results_lm7 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Kreatinine1) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Kreatinine1")

results_lm8 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ body_fat) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "body_fat")

results_lm9 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_sex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_sex")

results_lm <- rbind(results_lm1, results_lm2, results_lm3, results_lm4,results_lm5,results_lm6,results_lm7, results_lm8, results_lm9) |> 
  filter(term != "(Intercept)") |> 
  select(-std.error, -statistic) |> 
  mutate(abs_estimate = abs(estimate),
         direction = estimate / abs_estimate,
         p.adj = p.adjust(p.value, method = "fdr"),
         log10_p_value_and_direction = -log10(p.adj) * direction,
         FDR = cut(log10_p_value_and_direction, 
                   breaks =c(-35, -6, -4, -2.301, -2, -1.301, 1, 1.301, 2, 2.301, 4, 6, 35),
                   labels =c("<0.000001 (neg)", "<0.0001 (neg)", "<0.005 (neg)", "<0.01 (neg)", "<0.05 (neg)", "0", "0",
                             "<0.05 (pos)", "<0.01 (pos)", "<0.005 (pos)", "<0.0001 (pos)", "<0.000001 (pos)")))

# Creating bubble plot.
palette_heatmap <- c("<0.000001 (pos)"="#800026",
                     "<0.0001 (pos)"="#BD0026",
                     "<0.005 (pos)"="#E31A1C",
                     "<0.01 (pos)"="#FC4E2A",
                     "<0.05 (pos)"="#FD8D3C",
                     "non-signif (pos)"="khaki1",
                     "non-signif"="white", 
                     "non-signif (neg)"="lightcyan",
                     "<0.05 (neg)"="#6BAED6",
                     "<0.01 (neg)"="#4292C6",
                     "<0.005 (neg)"="#2171B5",
                     "<0.0001 (neg)"="#08519C",
                     "<0.000001 (neg)"="#08306B")


 results_lm <- results_lm |> 
  mutate(
    marker = factor(marker, levels = rev(unique(marker))),
    term = case_when(
      term == "anth_bmi" ~ "BMI",
      term == "dem_age" ~ "Age",
      term == "HbA1c_value" ~ "HbA1c",
      term == "dm_dur_in_years" ~ "DM duration",
      term == "dem_sex" ~ "Sex (Male)",
      term == "Kreatinine1" ~ "Creatinine",
      term == "visceralfat_lev" ~ "Visceral Fat",
      term == "CentralFatIndex" ~ "Central Fat",
      term == "body_fat" ~ "Total Fat",
      TRUE ~ term
    )
  )



# Plot bubble plot
library(ggplot2)

 p <-ggplot( data = results_lm, aes(x = term, 
             y = marker, 
             size = abs_estimate, 
             color = FDR)) + 
  geom_point(shape = 15) + 
  scale_size_continuous(range = c(1, 3.5), breaks = c(0.1, 0.5, 1 , 2, 3, 4, 5, 6,7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17 )) +
  scale_color_manual(values = palette_heatmap, na.value = "grey90") +
  labs(size = "Effect size\n(absolute)") + 
  #facet_grid(~ Phenotype, scale = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 0, size = 10), # changing metadata text size
        axis.text.y = element_text(size = 10),  # changing y axis text size (proteins)
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.3, "line"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(), 
        panel.background = element_rect(fill = "gainsboro"),
        panel.grid = element_blank(),      
        strip.text.y = element_text(size = 5), # changing facet grid text size
        strip.text.x = element_text(size = 10), # changing phenotype text size 
        strip.placement = "outside") + # for placing the strip outside of the axis names
  scale_x_discrete(position = "top")


ggsave("plotMBElipids.png", plot = p, width = 6, height = 30, units = "in", dpi=600)
```



By Sex: Male
```{r}


# Subset data for Mos site
dat_male <- subset(dat_de_master, dem_sex == "Male")

# Subset data for Mbe site
dat_female <- subset(dat_de_master, dem_sex == "Female")


dat_de_master$HbA1c_value

# Define predictor variables
predictors <- c("anth_bmi", "dem_age","HbA1c_value", "Site","dm_dur_in_years", "visceralfat_lev", "Kreatinine1","CentralFatIndex", "body_fat")

# Extract inflammatory marker column names (columns 63 to 139)
lipid_markers <- names(dat_male)[145:381]

# Combine all required variables
all_vars <- c(predictors, lipid_markers)

# Check if all variables exist in the dataset
missing <- setdiff(all_vars, names(dat_male))
if (length(missing) > 0) {
  stop("Missing variables: ", paste(missing, collapse = ", "))
}

# Subset and clean the data
df_sub <- dat_male [, all_vars]

df_sub$HbA1c_value[df_sub$HbA1c_value == -99] <- NA

# Example data frame
df_sub <- df_sub


# Convert to long format so we can group by outcome variable
df_long <- df_sub |> 
  pivot_longer(cols = all_of(lipid_markers), 
               names_to = "marker", 
               values_to = "NPX")

# Run regression grouped by each marker (like your `results_lm1`)
results_lm1 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ anth_bmi) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)  |>
    mutate(term = "anth_bmi")
  

results_lm2 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_age) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_age")

results_lm3 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dm_dur_in_years) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dm_dur_in_years")


results_lm4 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ HbA1c_value) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "HbA1c_value")

results_lm5 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ CentralFatIndex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "CentralFatIndex")

results_lm6 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~visceralfat_lev ) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "visceralfat_lev")

results_lm7 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Kreatinine1) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Kreatinine1")

results_lm8 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ body_fat) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "body_fat")


results_lm9 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Site) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Site")

results_lm <- rbind(results_lm1, results_lm2, results_lm3, results_lm4,results_lm5,results_lm6,results_lm7, results_lm8, results_lm9) |> 
  filter(term != "(Intercept)") |> 
  select(-std.error, -statistic) |> 
  mutate(abs_estimate = abs(estimate),
         direction = estimate / abs_estimate,
         p.adj = p.adjust(p.value, method = "fdr"),
         log10_p_value_and_direction = -log10(p.adj) * direction,
         FDR = cut(log10_p_value_and_direction, 
                   breaks =c(-35, -6, -4, -2.301, -2, -1.301, 1, 1.301, 2, 2.301, 4, 6, 35),
                   labels =c("<0.000001 (neg)", "<0.0001 (neg)", "<0.005 (neg)", "<0.01 (neg)", "<0.05 (neg)", "0", "0",
                             "<0.05 (pos)", "<0.01 (pos)", "<0.005 (pos)", "<0.0001 (pos)", "<0.000001 (pos)")))

# Creating bubble plot.
palette_heatmap <- c("<0.000001 (pos)"="#800026",
                     "<0.0001 (pos)"="#BD0026",
                     "<0.005 (pos)"="#E31A1C",
                     "<0.01 (pos)"="#FC4E2A",
                     "<0.05 (pos)"="#FD8D3C",
                     "non-signif (pos)"="khaki1",
                     "non-signif"="white", 
                     "non-signif (neg)"="lightcyan",
                     "<0.05 (neg)"="#6BAED6",
                     "<0.01 (neg)"="#4292C6",
                     "<0.005 (neg)"="#2171B5",
                     "<0.0001 (neg)"="#08519C",
                     "<0.000001 (neg)"="#08306B")


 results_lm <- results_lm |> 
  mutate(
    marker = factor(marker, levels = rev(unique(marker))),
    term = case_when(
      term == "anth_bmi" ~ "BMI",
      term == "dem_age" ~ "Age",
      term == "HbA1c_value" ~ "HbA1c",
      term == "dm_dur_in_years" ~ "DM duration",
      term == "Site" ~ "Site (Moshi)",
      term == "Kreatinine1" ~ "Creatinine",
      term == "visceralfat_lev" ~ "Visceral Fat",
      term == "CentralFatIndex" ~ "Central Fat",
      term == "body_fat" ~ "Total Fat",
      TRUE ~ term
    )
  )
 
 


# Plot bubble plot
library(ggplot2)

 p <-ggplot( data = results_lm, aes(x = term, 
             y = marker, 
             size = abs_estimate, 
             color = FDR)) + 
  geom_point(shape = 15) + 
  scale_size_continuous(range = c(1, 3.5), breaks = c(0.1, 0.5, 1 , 2, 3, 4, 5, 6,7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17 )) +
  scale_color_manual(values = palette_heatmap, na.value = "grey90") +
  labs(size = "Effect size\n(absolute)") + 
  #facet_grid(~ Phenotype, scale = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 0, size = 10), # changing metadata text size
        axis.text.y = element_text(size = 10),  # changing y axis text size (proteins)
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.3, "line"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(), 
        panel.background = element_rect(fill = "gainsboro"),
        panel.grid = element_blank(),      
        strip.text.y = element_text(size = 5), # changing facet grid text size
        strip.text.x = element_text(size = 10), # changing phenotype text size 
        strip.placement = "outside") + # for placing the strip outside of the axis names
  scale_x_discrete(position = "top")

ggsave("plotMalelipid.png", plot = p, width = 6, height = 30, units = "in")

```


By Sex: Female
```{r}


# Subset data for Mos site
dat_male <- subset(dat_de_master, dem_sex == "Male")
colnames(dat_male)
# Subset data for Mbe site
dat_female <- subset(dat_de_master, dem_sex == "Female")


# Define predictor variables
predictors <- c("anth_bmi", "dem_age","HbA1c_value", "Site","dm_dur_in_years", "visceralfat_lev", "Kreatinine1","CentralFatIndex", "body_fat")

# Extract inflammatory marker column names (columns 63 to 139)
lipid_markers <- names(dat_male)[145:381]

# Combine all required variables
all_vars <- c(predictors, lipid_markers)

# Check if all variables exist in the dataset
missing <- setdiff(all_vars, names(dat_male))
if (length(missing) > 0) {
  stop("Missing variables: ", paste(missing, collapse = ", "))
}

# Subset and clean the data
df_sub <- dat_female [, all_vars]

df_sub$HbA1c_value[df_sub$HbA1c_value == -99] <- NA

# Example data frame
df_sub <- df_sub

library()
# Convert to long format so we can group by outcome variable
library(dplyr)
library(tidyr)

df_long <- df_sub |> 
  pivot_longer(cols = all_of(lipid_markers), 
               names_to = "marker", 
               values_to = "NPX")

# Run regression grouped by each marker (like your `results_lm1`)
results_lm1 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ anth_bmi) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)  |>
    mutate(term = "anth_bmi")
  

results_lm2 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dem_age) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dem_age")

results_lm3 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ dm_dur_in_years) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "dm_dur_in_years")


results_lm4 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ HbA1c_value) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "HbA1c_value")

results_lm5 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ CentralFatIndex) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "CentralFatIndex")

results_lm6 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~visceralfat_lev ) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "visceralfat_lev")

results_lm7 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Kreatinine1) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Kreatinine1")

results_lm8 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ body_fat) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "body_fat")


results_lm9 <- df_long |> 
  group_by(marker) |> 
  summarize(
    out = list(lm(NPX ~ Site) |> broom::tidy()),
    .groups = "drop"
  ) |> 
  unnest(out)|> 
 mutate(term = "Site")

results_lm <- rbind(results_lm1, results_lm2, results_lm3, results_lm4,results_lm5,results_lm6,results_lm7, results_lm8, results_lm9) |> 
  filter(term != "(Intercept)") |> 
  select(-std.error, -statistic) |> 
  mutate(abs_estimate = abs(estimate),
         direction = estimate / abs_estimate,
         p.adj = p.adjust(p.value, method = "fdr"),
         log10_p_value_and_direction = -log10(p.adj) * direction,
         FDR = cut(log10_p_value_and_direction, 
                   breaks =c(-35, -6, -4, -2.301, -2, -1.301, 1, 1.301, 2, 2.301, 4, 6, 35),
                   labels =c("<0.000001 (neg)", "<0.0001 (neg)", "<0.005 (neg)", "<0.01 (neg)", "<0.05 (neg)", "0", "0",
                             "<0.05 (pos)", "<0.01 (pos)", "<0.005 (pos)", "<0.0001 (pos)", "<0.000001 (pos)")))

# Creating bubble plot.
palette_heatmap <- c("<0.000001 (pos)"="#800026",
                     "<0.0001 (pos)"="#BD0026",
                     "<0.005 (pos)"="#E31A1C",
                     "<0.01 (pos)"="#FC4E2A",
                     "<0.05 (pos)"="#FD8D3C",
                     "non-signif (pos)"="khaki1",
                     "non-signif"="white", 
                     "non-signif (neg)"="lightcyan",
                     "<0.05 (neg)"="#6BAED6",
                     "<0.01 (neg)"="#4292C6",
                     "<0.005 (neg)"="#2171B5",
                     "<0.0001 (neg)"="#08519C",
                     "<0.000001 (neg)"="#08306B")


 results_lm <- results_lm |> 
  mutate(
    marker = factor(marker, levels = rev(unique(marker))),
    term = case_when(
      term == "anth_bmi" ~ "BMI",
      term == "dem_age" ~ "Age",
      term == "HbA1c_value" ~ "HbA1c",
      term == "dm_dur_in_years" ~ "DM duration",
      term == "Site" ~ "Site (Moshi)",
      term == "Kreatinine1" ~ "Creatinine",
      term == "visceralfat_lev" ~ "Visceral Fat",
      term == "CentralFatIndex" ~ "Central Fat",
      term == "body_fat" ~ "Total Fat",
      TRUE ~ term
    )
  )


# Plot bubble plot
library(ggplot2)

 p <-ggplot( data = results_lm, aes(x = term, 
             y = marker, 
             size = abs_estimate, 
             color = FDR)) + 
  geom_point(shape = 15) + 
  scale_size_continuous(range = c(1, 3.5), breaks = c(0.1, 0.5, 1 , 2, 3, 4, 5, 6,7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17 )) +
  scale_color_manual(values = palette_heatmap, na.value = "grey90") +
  labs(size = "Effect size\n(absolute)") + 
  #facet_grid(~ Phenotype, scale = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 0, size = 10), # changing metadata text size
        axis.text.y = element_text(size = 10),  # changing y axis text size (proteins)
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.3, "line"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(), 
        panel.background = element_rect(fill = "gainsboro"),
        panel.grid = element_blank(),      
        strip.text.y = element_text(size = 5), # changing facet grid text size
        strip.text.x = element_text(size = 10), # changing phenotype text size 
        strip.placement = "outside") + # for placing the strip outside of the axis names
  scale_x_discrete(position = "top")


ggsave("plotfemalelipids.png", plot = p, width = 6, height = 28, units = "in")
```

#Comparison between endotypesA and C (Olink data only)
```{r}
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(ggrepel)

# Define the comparison
comparison <- "Endotype A vs Endotype C"

# Select only the biomarker columns and cluster column
biomarker_data <- dat_de_master_lipid |> 
  select(63:139, Cluster = 140) |>   # rename column 140 as 'Cluster'
  filter(Cluster %in% c("Endotype A", "Endotype C"))

colnames(biomarker_data)

# Convert to long format for easier summarization
long_data <- biomarker_data |> 
  pivot_longer(
    cols = 1:77,          # 63:139 = 77 columns
    names_to = "Biomarker",
    values_to = "Concentration"
  ) |> 
  filter(!is.na(Concentration))

# Summarize mean concentration per biomarker and cluster
marker_summary <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(mean_conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = mean_conc) |>
  mutate(
    diff = `Endotype A` - `Endotype C`,
    regulation = case_when(
      diff > 0 ~ "Up in Endotype A",
      diff < 0 ~ "Up in Endotype C",
      TRUE ~ "No difference"
    )
  )

# View results
marker_summary


# 2. Calculate log2FC
log2FC_df <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = Mean_Conc) |>
  mutate(log2FC = log2(`Endotype C` / `Endotype A`))

# 3. Wilcoxon rank-sum test for each biomarker
p_values <- long_data |>
  group_by(Biomarker) |>
  summarise(out = list(wilcox.test(Concentration ~ Cluster, paired = FALSE) |> broom::tidy()), .groups = "drop") |>
  unnest(cols = c(out))

# 4. Combine log2FC and p-values
tmp <- log2FC_df |>
  left_join(p_values, by = "Biomarker") |>
   ungroup() |>
  mutate(
    p.adj = p.adjust(p.value, method = "fdr"),
    `-log10P` = -log10(p.adj),
    signif = ifelse(p.adj < 0.05, "Significant", "Non-significant")
  )

labels <- tmp |> 
  filter(p.adj < 0.05)


# . Define plot limits
max_l2FC <- max(abs(tmp$log2FC), na.rm = TRUE)
min_p <- min(tmp$p.adj, na.rm = TRUE)

# plot with making labels infinite and thinner
tmp |> 
  ggplot(aes(x = log2FC, y = `-log10P`)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = -log10(0.05), alpha = 0.5, color = "darkred", linetype = 2) +
  geom_point(aes(color = signif), show.legend = F) +
  ggrepel::geom_text_repel(data = subset(tmp, signif == "Significant"), # Only label significant dots
                           aes(label = Biomarker), 
                           size = 2.5,           # Reduce label size
                           max.overlaps = Inf,   # Allow all overlaps for significant dots
                           nudge_x = 0.1,        # Adjust x position for better spacing
                           nudge_y = 0.1,        # Adjust y position for better spacing
                           segment.size = 0.2) + # Make connecting lines thinner
  scale_y_continuous(limits = c(0, -log10(min_p)), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(limits = c(-max_l2FC, max_l2FC)) +
  scale_color_discrete(limits = c("Significant", "Non-significant"), type = c("red2", "gray")) +
  labs(title = comparison,
       y = "-log10(adjusted p-value)",
       x = "log2FC") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        panel.grid = element_blank())

# Save figure
ggsave(
  filename = "Olink_EndotypeC_vs_EndotypeA.png",
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4,
  width = 4,
  units = "in"
)

```

Endotype A vs B 

```{r}
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(ggrepel)

# Define the comparison
comparison <- "Endotype A vs Endotype B"

# Select only the biomarker columns and cluster column
biomarker_data <- dat_de_master_lipid |> 
  select(63:139, Cluster = 140) |>   # rename column 140 as 'Cluster'
  filter(Cluster %in% c("Endotype A", "Endotype B"))

colnames(biomarker_data)

# Convert to long format for easier summarization
long_data <- biomarker_data |> 
  pivot_longer(
    cols = 1:77,          # 63:139 = 77 columns
    names_to = "Biomarker",
    values_to = "Concentration"
  ) |> 
  filter(!is.na(Concentration))

# Summarize mean concentration per biomarker and cluster
marker_summary <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(mean_conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = mean_conc) |>
  mutate(
    diff = `Endotype A` - `Endotype B`,
    regulation = case_when(
      diff > 0 ~ "Up in Endotype A",
      diff < 0 ~ "Up in Endotype B",
      TRUE ~ "No difference"
    )
  )

# View results
marker_summary


# 2. Calculate log2FC
log2FC_df <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = Mean_Conc) |>
  mutate(log2FC = log2(`Endotype B` / `Endotype A`))

# 3. Wilcoxon rank-sum test for each biomarker
p_values <- long_data |>
  group_by(Biomarker) |>
  summarise(out = list(wilcox.test(Concentration ~ Cluster, paired = FALSE) |> broom::tidy()), .groups = "drop") |>
  unnest(cols = c(out))

# 4. Combine log2FC and p-values
tmp <- log2FC_df |>
  left_join(p_values, by = "Biomarker") |>
   ungroup() |>
  mutate(
    p.adj = p.adjust(p.value, method = "fdr"),
    `-log10P` = -log10(p.adj),
    signif = ifelse(p.adj < 0.05, "Significant", "Non-significant")
  )

labels <- tmp |> 
  filter(p.adj < 0.05)


# . Define plot limits
max_l2FC <- max(abs(tmp$log2FC), na.rm = TRUE)
min_p <- min(tmp$p.adj, na.rm = TRUE)

# plot with making labels infinite and thinner
tmp |> 
  ggplot(aes(x = log2FC, y = `-log10P`)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = -log10(0.05), alpha = 0.5, color = "darkred", linetype = 2) +
  geom_point(aes(color = signif), show.legend = F) +
  ggrepel::geom_text_repel(data = subset(tmp, signif == "Significant"), # Only label significant dots
                           aes(label = Biomarker), 
                           size = 2.5,           # Reduce label size
                           max.overlaps = Inf,   # Allow all overlaps for significant dots
                           nudge_x = 0.1,        # Adjust x position for better spacing
                           nudge_y = 0.1,        # Adjust y position for better spacing
                           segment.size = 0.2) + # Make connecting lines thinner
  scale_y_continuous(limits = c(0, -log10(min_p)), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(limits = c(-max_l2FC, max_l2FC)) +
  scale_color_discrete(limits = c("Significant", "Non-significant"), type = c("red2", "gray")) +
  labs(title = comparison,
       y = "-log10(adjusted p-value)",
       x = "log2FC") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        panel.grid = element_blank())

# Save figure
ggsave(
  filename = "Olink_EndotypeA_vs_EndotypeB.png",
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4,
  width = 4,
  units = "in"
)
```
Endotype B VS C 
```{r}
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(ggrepel)

# Define the comparison
comparison <- "Endotype B vs Endotype C"

# Select only the biomarker columns and cluster column
biomarker_data <- dat_de_master_lipid |> 
  select(63:139, Cluster = 140) |>   # rename column 140 as 'Cluster'
  filter(Cluster %in% c("Endotype B", "Endotype C"))

colnames(biomarker_data)

# Convert to long format for easier summarization
long_data <- biomarker_data |> 
  pivot_longer(
    cols = 1:77,          # 63:139 = 77 columns
    names_to = "Biomarker",
    values_to = "Concentration"
  ) |> 
  filter(!is.na(Concentration))

# Summarize mean concentration per biomarker and cluster
marker_summary <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(mean_conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = mean_conc) |>
  mutate(
    diff = `Endotype B` - `Endotype C`,
    regulation = case_when(
      diff > 0 ~ "Up in Endotype B",
      diff < 0 ~ "Up in Endotype C",
      TRUE ~ "No difference"
    )
  )

# View results
marker_summary


# 2. Calculate log2FC
log2FC_df <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Cluster, values_from = Mean_Conc) |>
  mutate(log2FC = log2(`Endotype B` / `Endotype C`))

# 3. Wilcoxon rank-sum test for each biomarker
p_values <- long_data |>
  group_by(Biomarker) |>
  summarise(out = list(wilcox.test(Concentration ~ Cluster, paired = FALSE) |> broom::tidy()), .groups = "drop") |>
  unnest(cols = c(out))

# 4. Combine log2FC and p-values
tmp <- log2FC_df |>
  left_join(p_values, by = "Biomarker") |>
   ungroup() |>
  mutate(
    p.adj = p.adjust(p.value, method = "fdr"),
    `-log10P` = -log10(p.adj),
    signif = ifelse(p.adj < 0.05, "Significant", "Non-significant")
  )

labels <- tmp |> 
  filter(p.adj < 0.05)


# . Define plot limits
max_l2FC <- max(abs(tmp$log2FC), na.rm = TRUE)
min_p <- min(tmp$p.adj, na.rm = TRUE)

# plot with making labels infinite and thinner
tmp |> 
  ggplot(aes(x = log2FC, y = `-log10P`)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = -log10(0.05), alpha = 0.5, color = "darkred", linetype = 2) +
  geom_point(aes(color = signif), show.legend = F) +
  ggrepel::geom_text_repel(data = subset(tmp, signif == "Significant"), # Only label significant dots
                           aes(label = Biomarker), 
                           size = 2.5,           # Reduce label size
                           max.overlaps = Inf,   # Allow all overlaps for significant dots
                           nudge_x = 0.1,        # Adjust x position for better spacing
                           nudge_y = 0.1,        # Adjust y position for better spacing
                           segment.size = 0.2) + # Make connecting lines thinner
  scale_y_continuous(limits = c(0, -log10(min_p)), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(limits = c(-max_l2FC, max_l2FC)) +
  scale_color_discrete(limits = c("Significant", "Non-significant"), type = c("red2", "gray")) +
  labs(title = comparison,
       y = "-log10(adjusted p-value)",
       x = "log2FC") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        panel.grid = element_blank())

# Save figure
ggsave(
  filename = "Olink_EndotypeB_vs_EndotypeC.png",
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4,
  width = 4,
  units = "in"
)
```
Combined plots
```{r}
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(ggrepel)

install.packages("patchwork")
library(patchwork)   # for combining plots

# Function to generate volcano plot for any two clusters
make_volcano <- function(data, group1, group2, title) {
  long_data <- data |>
    filter(Cluster %in% c(group1, group2)) |>
    pivot_longer(
      cols = 1:77,  # biomarker columns
      names_to = "Biomarker",
      values_to = "Concentration"
    ) |> 
    filter(!is.na(Concentration))
  
  # Mean values
  log2FC_df <- long_data |>
    group_by(Biomarker, Cluster) |>
    summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = Cluster, values_from = Mean_Conc) |>
    mutate(log2FC = log2(.data[[group2]] / .data[[group1]]))
  
  # Wilcoxon test
  p_values <- long_data |>
    group_by(Biomarker) |>
    summarise(out = list(wilcox.test(Concentration ~ Cluster, paired = FALSE) |> broom::tidy()),
              .groups = "drop") |>
    unnest(cols = c(out))
  
  # Merge results
  tmp <- log2FC_df |>
    left_join(p_values, by = "Biomarker") |>
    ungroup() |>
    mutate(
      p.adj = p.adjust(p.value, method = "fdr"),
      `-log10P` = -log10(p.adj),
      signif = ifelse(p.adj < 0.05, "Significant", "Non-significant")
    )
  
  # Volcano plot
  max_l2FC <- max(abs(tmp$log2FC), na.rm = TRUE)
  
  ggplot(tmp, aes(x = log2FC, y = `-log10P`)) +
    geom_vline(xintercept = 0, color = "black") +
    geom_hline(yintercept = -log10(0.05), color = "darkred", linetype = 2, alpha = 0.6) +
    geom_point(aes(color = signif), size = 2, alpha = 0.7, show.legend = FALSE) +
    ggrepel::geom_text_repel(
      data = subset(tmp, signif == "Significant"),
      aes(label = Biomarker),
      size = 2.5, max.overlaps = Inf, segment.size = 0.2
    ) +
    scale_x_continuous(limits = c(-max_l2FC, max_l2FC)) +
    scale_color_manual(values = c("Significant" = "red2", "Non-significant" = "gray70")) +
    labs(title = title, x = "log2FC", y = "-log10(adjusted p-value)") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), panel.grid = element_blank())
}

# Prepare input data
biomarker_data <- dat_de_master_lipid |>
  select(63:139, Cluster = 140)   # biomarkers + cluster

# Generate plots
p1 <- make_volcano(biomarker_data, "Endotype A", "Endotype B", "Endotype A vs B")
p2 <- make_volcano(biomarker_data, "Endotype B", "Endotype C", "Endotype B vs C")
p3 <- make_volcano(biomarker_data, "Endotype A", "Endotype C", "Endotype A vs C")

# Combine plots side by side (or in 2x2 grid if you had more)
combined <- (p1 | p2 | p3)

# Save combined figure
ggsave(
  "Olink_EndotypeComparisons.png",
  combined,
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4, width = 12, units = "in"
)

```




```{r}
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(ggrepel)

# Function to calculate stats for any pair
do_comparison <- function(data, g1, g2) {
  long_data <- data |>
    filter(Cluster %in% c(g1, g2)) |>
    pivot_longer(
      cols = 1:77, 
      names_to = "Biomarker",
      values_to = "Concentration"
    ) |> 
    filter(!is.na(Concentration))
  
  # Mean values
  log2FC_df <- long_data |>
    group_by(Biomarker, Cluster) |>
    summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = Cluster, values_from = Mean_Conc) |>
    mutate(log2FC = log2(.data[[g2]] / .data[[g1]]))
  
  # Wilcoxon test
  p_values <- long_data |>
    group_by(Biomarker) |>
    summarise(out = list(wilcox.test(Concentration ~ Cluster, paired = FALSE) |> broom::tidy()),
              .groups = "drop") |>
    unnest(cols = c(out))
  
  # Merge results
  res <- log2FC_df |>
    left_join(p_values, by = "Biomarker") |>
    ungroup() |>
    mutate(
      p.adj = p.adjust(p.value, method = "fdr"),
      `-log10P` = -log10(p.adj),
      signif = ifelse(p.adj < 0.05, "Significant", "Non-significant"),
      Comparison = paste(g1, "vs", g2)
    )
  
  return(res)
}

# Prepare input data
biomarker_data <- dat_de_master_lipid |>
  select(63:139, Cluster = 140)

# Run all three comparisons
res_AB <- do_comparison(biomarker_data, "Endotype A", "Endotype B")
res_BC <- do_comparison(biomarker_data, "Endotype B", "Endotype C")
res_AC <- do_comparison(biomarker_data, "Endotype A", "Endotype C")

# Combine into one dataset
all_res <- bind_rows(res_AB, res_BC, res_AC)

# Volcano plot with facets for each comparison
ggplot(all_res, aes(x = log2FC, y = `-log10P`)) +
  geom_vline(xintercept = 0, color = "black") +
  geom_hline(yintercept = -log10(0.05), color = "darkred", linetype = 2, alpha = 0.6) +
  geom_point(aes(color = signif), size = 1.8, alpha = 0.7) +
  ggrepel::geom_text_repel(
    data = subset(all_res, signif == "Significant"),
    aes(label = Biomarker),
    size = 2,
    max.overlaps = 20,
    segment.size = 0.2
  ) +
  scale_color_manual(values = c("Significant" = "red2", "Non-significant" = "gray70")) +
  labs(
    title = "Volcano plot of Endotype Comparisons",
    x = "log2FC",
    y = "-log10(adjusted p-value)"
  ) +
  facet_wrap(~Comparison, ncol = 3) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    legend.title = element_blank()
  )

# Save figure
ggsave(
  "Olink_All_EndotypeComparisons.png",
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4, width = 12, units = "in"
)

```


### Radial plot for Olink endotypes 
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(volcano3D) 

#--------------------------------------
# Step 1: Set phenotype/endotype order
#--------------------------------------
# Your biomarker dataset: columns 63–139 = biomarkers, 140 = cluster
biomarker_data <- dat_de_master|> 
  select(63:139, Cluster = 140)

# Step 1: Select biomarkers + Cluster
biomarker_data <- dat_de_master |> 
  select(63:139, Cluster = 140)

# Ensure Cluster is an ordered factor
biomarker_data$Cluster <- factor(
  biomarker_data$Cluster,
  levels = c("Endotype A", "Endotype B", "Endotype C"),
  ordered = TRUE
)

# Step 2: Loop over biomarkers in wide format
library(dplyr)
library(tidyr)

biomarker_names <- colnames(biomarker_data)[1:77]  # adjust 63:139 = 77 columns

kruskal_results <- lapply(biomarker_names, function(gene){
  df <- biomarker_data |> select(Cluster, Value = all_of(gene))
  
  # Kruskal-Wallis test
  kw <- kruskal.test(Value ~ Cluster, data = df)$p.value
  

 # Pairwise Wilcoxon (reference df$Cluster and df$Value)
  pw_A_B <- wilcox.test(df$Value[df$Cluster == "Endotype A"],
                        df$Value[df$Cluster == "Endotype B"])$p.value
  pw_A_C <- wilcox.test(df$Value[df$Cluster == "Endotype A"],
                        df$Value[df$Cluster == "Endotype C"])$p.value
  pw_B_C <- wilcox.test(df$Value[df$Cluster == "Endotype B"],
                        df$Value[df$Cluster == "Endotype C"])$p.value
  
  tibble(
    Biomarker = gene,
    Kruskal_result = kw,
    p.adj_A_vs_B = pw_A_B,
    p.adj_A_vs_C = pw_A_C,
    p.adj_B_vs_C = pw_B_C
  )
}) |> bind_rows() |> 
  mutate(
    Kruskal_result = p.adjust(Kruskal_result, method = "BH"),
    p.adj_A_vs_B = p.adjust(p.adj_A_vs_B, method = "BH"),
    p.adj_A_vs_C = p.adjust(p.adj_A_vs_C, method = "BH"),
    p.adj_B_vs_C = p.adjust(p.adj_B_vs_C, method = "BH")
  )

kruskal_results

# Step 3: Prepare matrix for volcano plotting
#--------------------------------------
tmp <- kruskal_results |> 
  column_to_rownames("Biomarker") |> 
  as.matrix()


# Add rownames as a column
biomarker_data <- biomarker_data |> 
  tibble::rownames_to_column(var = "SampleID")

# Now create test_metadata
test_metadata <- biomarker_data |> 
  select(SampleID, Cluster) |> 
  unique() |> 
  mutate(Cluster = factor(Cluster, levels = c("Endotype A", "Endotype B", "Endotype C")))

test_matrix <- biomarker_data |> 
  select(SampleID, all_of(biomarker_names), Cluster) |>  # keep SampleID, biomarkers, Cluster
  column_to_rownames("SampleID") |>                      # set SampleID as rownames
  select(-Cluster) |>                                     # remove Cluster (only numeric values)
  as.matrix()                                            # convert to numeric matrix


# Ensure order of biomarkers in tmp matches columns in test_matrix
tmp1 <- tmp[colnames(test_matrix),]


#--------------------------------------
# Step 5: Generate polar coordinates for volcano
#--------------------------------------
test_polar <- polar_coords(outcome = test_metadata$Cluster,
                            data = test_matrix,
                            pvals = tmp1)



#--------------------------------------
# Step 6: 3D Volcano plot (optional)
#--------------------------------------
 volcano3D(test_polar)

#--------------------------------------
# Step 7: 2D Volcano plot
#--------------------------------------
tmp_plot <- radial_ggplot(polar = test_polar)
 tmp_plot$data <- tmp_plot$data |>
   rownames_to_column("Gene") |>
   mutate(Gene = case_when(lab != "ns" ~ Gene),
          lab = factor(lab, levels = c("ns", "Endotype A", "Endotype B", "Endotype C")))

tmp_plot + 
   ggrepel::geom_text_repel(aes(label = Gene), size = 3.5,  max.overlaps = 100) +
   theme(legend.text = element_text(size = 12),
         legend.position = "bottom")


all(colnames(test_matrix) == rownames(tmp1))

levels(test_metadata$Cluster)

write.csv(tmp1, "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/tmp1_results.csv", row.names = TRUE)



```

```{r}
library(ggrepel)

p <- ggplot(mean_data, aes(x = angle_rad, y = Mean_Conc)) +
  geom_vline(data = quadrant_lines, aes(xintercept = angle_rad),
             color = "black", linetype = "dashed") +
  geom_point(aes(color = color_group), shape = 19, size = 4, alpha = 0.8) +
  geom_text_repel(aes(label = Biomarker),
                  size = 3,
                  min.segment.length = 0,
                  segment.color = "gray50") +
  geom_text(data = endotype_labels,
            aes(x = angle_rad, y = radius, label = Cluster),
            inherit.aes = FALSE,
            size = 4, fontface = "bold") +
  scale_color_manual(values = c(
    "Endotype A" = "#2ca02c",
    "Endotype B" = "#ff7f0e",
    "Endotype C" = "#d62728",
    "Non-significant" = "gray70"
  )) +
  coord_polar(theta = "x", start = 0) +
  labs(title = "Radial Quadrant Plot of Biomarkers by Endotype",
       x = "", y = "Mean Concentration") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_line(color = "gray80"),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    legend.position = "bottom"
  )

print(p)

```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)

# -----------------------------
# 1. Prepare data
# -----------------------------
long_data <- dat_de_master_lipid |>
  pivot_longer(
    cols = 63:139,
    names_to = "Biomarker",
    values_to = "Concentration"
  ) |>
  filter(!is.na(Concentration))

# Kruskal-Wallis test per biomarker
kruskal_res <- long_data |>
  group_by(Biomarker) |>
  summarise(
    test = list(kruskal.test(Concentration ~ Cluster) %>% broom::tidy()),
    .groups = "drop"
  ) |>
  unnest(cols = c(test)) |>
  mutate(
    p.adj = p.adjust(p.value, method = "fdr"),
    signif = ifelse(p.adj < 0.05, "Significant", "Non-significant")
  )

# Mean concentration per biomarker & endotype
mean_data <- long_data |>
  group_by(Biomarker, Cluster) |>
  summarise(Mean_Conc = mean(Concentration, na.rm = TRUE), .groups = "drop") |>
  left_join(kruskal_res |> select(Biomarker, signif), by = "Biomarker")

# -----------------------------
# 2. Assign angles for radial layout
# -----------------------------
mean_data <- mean_data |>
  group_by(Cluster) |>
  mutate(
    idx = row_number(),
    angle = case_when(
      Cluster == "Endotype A" ~ 0 + (idx - 1) * (120 / n()),
      Cluster == "Endotype B" ~ 120 + (idx - 1) * (120 / n()),
      Cluster == "Endotype C" ~ 240 + (idx - 1) * (120 / n())
    ),
    angle_rad = angle * pi / 180,
    color_group = factor(
      ifelse(signif == "Significant", as.character(Cluster), "Non-significant"),
      levels = c("Endotype A", "Endotype B", "Endotype C", "Non-significant")
    )
  ) |>
  ungroup()

# -----------------------------
# 3. Quadrant lines and labels
# -----------------------------
quadrant_lines <- data.frame(angle_rad = c(0, 120, 240) * pi / 180)

endotype_labels <- data.frame(
  Cluster = c("Endotype A", "Endotype B", "Endotype C"),
  angle = c(60, 180, 300),
  radius = max(mean_data$Mean_Conc) + 0.2
) |>
  mutate(angle_rad = angle * pi / 180)

# -----------------------------
# 4. Radial plot
# -----------------------------
p <- ggplot(mean_data, aes(x = angle_rad, y = Mean_Conc)) +
  geom_vline(data = quadrant_lines, aes(xintercept = angle_rad),
             color = "black", linetype = "dashed") +
  geom_point(aes(color = color_group), shape = 19, size = 4, alpha = 0.8) +
  geom_text(aes(label = Biomarker), color = "black", size = 3, vjust = -0.5, hjust = 0.5) +
  geom_text(data = endotype_labels,
            aes(x = angle_rad, y = radius, label = Cluster),
            inherit.aes = FALSE,
            size = 4, fontface = "bold") +
  scale_color_manual(values = c(
    "Endotype A" = "#2ca02c",
    "Endotype B" = "#ff7f0e",
    "Endotype C" = "#d62728",
    "Non-significant" = "gray70"
  )) +
  coord_polar(theta = "x", start = 0) +
  labs(
    title = "Radial Quadrant Plot of Biomarkers by Endotype",
    x = "", y = "Mean Concentration"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_line(color = "gray80"),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    legend.position = "bottom"
  )

# Print plot
print(p)

# -----------------------------
# 5. Save figure
# -----------------------------
ggsave(
  "Olink_All_EndotypeComparisons_radial_krus.png",
  plot = p,
  path = "C:/Users/lmrema/Documents/Local Disk D/Trainings/PhD/Analysis/Lipidomics/Lipidomics/Figures",
  height = 4, width = 12, units = "in"
)
```

Correlations Olink Vs Nightingale
```{r}
library(dplyr)

# Olink: wide format, one row per PID
olink_cols <- dat_de_master_lipid |> 
  select(c(1, 63: 139)) 

# Lipids: select PID + required columns
lipid_cols <- dat_de_master_lipid |> 
  select(c(1, 145:159, 179, 181:187, 189:195, 205:225))   # includes PID in column 1

merged_df <- inner_join(lipid_cols, olink_cols, by = "PID")

lipid_vars <- colnames(lipid_cols)[-1]  # all lipid variables except PID
olink_vars <- setdiff(colnames(olink_cols), "PID")  # all olink proteins

merged_df <- merged_df |>
  mutate(across(all_of(c(lipid_vars, olink_vars)), ~ as.numeric(unlist(.x))))

library(psych)
library(corrplot)

# Compute correlations: lipids (X) vs Olink (Y)
cor_res <- psych::corr.test(
  merged_df[, lipid_vars],
  merged_df[, olink_vars],
  method = "spearman",
  adjust = "fdr",
  use = "pairwise.complete.obs"
)

# Extract matrices
r_mat <- cor_res$r
p_matadj <- cor_res$p

library(pheatmap)

col_fun <- colorRampPalette(c("blue4", "white", "red4"))
# Define color palette for correlations

png("Correlation_Lipids_vs_Olink.png", width = 2000, height = 1800, res = 300)

corrplot(as.matrix(r_mat),
         p.mat = as.matrix(p_mat),
         method = "color",
         col = col_fun(200),
         insig = "label_sig",
         sig.level = c(0.0001, 0.001, 0.05),
         pch.cex = 0.7,
         pch.col = "black",
         tl.col = "black",      # text label color
         tl.cex = 0.5,          # reduce axis font size
         cl.cex = 0.7,          # reduce color legend font size
         order = "hclust",
         hclust.method = "complete",
         addrect = "",
         main = "Spearman correlation (Lipids vs Olink proteins)")

dev.off()

```

Unadjusted 
```{r}

library(psych)

cor_res_unadj <- psych::corr.test(
  x = merged_df[, lipid_vars],
  y = merged_df[, olink_vars],
  method = "spearman",
  adjust = "none",                  # 🔴 correct way to specify unadjusted
  use = "pairwise.complete.obs"
)

# Extract matrices
r_mat <- cor_res_unadj$r            # correlation coefficients
p_mat_unadj <- cor_res_unadj$p            # RAW (unadjusted) p-values


# ============================
# 2️⃣ Plot and save correlation heatmap
# ============================

png(
  filename = "Correlation_Lipids_vs_Olink_UNADJUSTED.png",
  width = 2000,
  height = 1800,
  res = 300
)

corrplot(
  r_mat,
  p.mat = p_mat,
  method = "color",
  col = col_fun(200),
  insig = "label_sig",               # mark insignificant correlations
  sig.level = c(0.0001, 0.001, 0.05),
  pch.cex = 0.7,
  pch.col = "black",
  tl.col = "black",
  tl.cex = 0.5,                      # axis label size
  cl.cex = 0.7,                      # color legend size
  order = "hclust",
  hclust.method = "complete",
  main = "Spearman correlation between lipids and Olink proteins\n(unadjusted p-values)"
)

dev.off()

p_adj_mat <- matrix(
  p.adjust(as.vector(p_mat), method = "BH"),
  nrow = nrow(p_mat),
  ncol = ncol(p_mat),
  dimnames = dimnames(p_mat)
)

# how many significant before vs after?
sum(p_matadj < 0.05, na.rm = TRUE)
sum(p_mat_unadj < 0.05, na.rm = TRUE)

plot(
  as.vector(p_matadj),
  as.vector(p_mat_unadj),
  pch = 16,
  cex = 0.4,
  xlab = "Raw p-values",
  ylab = "FDR-adjusted p-values",
  main = "Effect of FDR correction"
)
abline(0, 1, col = "red", lty = 2)



```

Correlation plot Endotypes plus selected lipids. 
```{r}
library(dplyr)
library(psych)
library(pheatmap)

# ---- ensure Cluster labels are correct ----
unique(dat_de_master_lipid$Cluster)

# ---- create analysis data frame ----
df <- dat_de_master_lipid %>%
  select(PID, Cluster, all_of(lipid_vars)) %>%
  mutate(across(all_of(lipid_vars), as.numeric)) %>%
  mutate(
    Endotype_A = as.numeric(Cluster == "Endotype A"),
    Endotype_B = as.numeric(Cluster == "Endotype B"),
    Endotype_C = as.numeric(Cluster == "Endotype C")
  )

endotype_vars <- c("Endotype_A", "Endotype_B", "Endotype_C")

# ---- correlation ----
cor_res <- psych::corr.test(
  df[, lipid_vars],
  df[, endotype_vars],
  method = "spearman",
  adjust = "fdr",
  use = "pairwise.complete.obs"
)

# ---- extract matrices ----
r_mat <- cor_res$r
p_mat <- cor_res$p
sig_mat <- ifelse(p_mat < 0.05, "*", "")

# ---- significance stars (*, **, ***) ----
sig_mat <- matrix("", nrow = nrow(p_mat), ncol = ncol(p_mat))
rownames(sig_mat) <- rownames(p_mat)
colnames(sig_mat) <- colnames(p_mat)

sig_mat[p_mat < 0.05]  <- "*"
sig_mat[p_mat < 0.01]  <- "**"
sig_mat[p_mat < 0.001] <- "***"

# ---- nicer column labels ----
colnames(r_mat) <- gsub("_", " ", colnames(r_mat))
colnames(sig_mat) <- colnames(r_mat)

# ---- save heatmap ----
png(
  filename = "Lipids_vs_Endotypes_A_B_C.png",
  width = 1000,
  height = 2400,
  res = 300
)

pheatmap(
  r_mat,
  color = colorRampPalette(c("blue4", "white", "red4"))(200),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = sig_mat,
  number_color = "black",
  fontsize_row = 6,
  fontsize_col = 10,
  main = ""
)

dev.off()


```
Volcano plot for differences between Male and female Olink
```{r}
library(dplyr)
library(ggplot2)

# Olink columns
olink_cols <- colnames(dat_de_master)[63:139]

results <- lapply(olink_cols, function(prot) {

  df <- dat_de_master %>%
    select(dem_sex, all_of(prot)) %>%
    filter(!is.na(dem_sex), !is.na(.data[[prot]]))

  # ensure two groups
  if (n_distinct(df$dem_sex) < 2) return(NULL)

  # Wilcoxon test
  pval <- wilcox.test(df[[prot]] ~ df$dem_sex)$p.value

  # Effect size: log2 median difference
med_male <- median(df[[prot]][df$dem_sex == "Male"], na.rm = TRUE)
med_fem  <- median(df[[prot]][df$dem_sex == "Female"], na.rm = TRUE)

  log2FC <- log2(med_male / med_fem)

  data.frame(
    protein = prot,
    log2FC = log2FC,
    pvalue = pval
  )
})

results_df <- bind_rows(results) %>%
  mutate(
    padj = p.adjust(pvalue, method = "BH"),
    significance = ifelse(padj < 0.05, "Significant", "Not significant"),
    neglog10_padj = -log10(padj)
  )

library(ggrepel)


ggplot(results_df, aes(x = log2FC, y = neglog10_padj)) +
  geom_point(aes(color = significance), size = 2, alpha = 0.8) +
  geom_text_repel(
    data = subset(results_df, padj < 0.05),
    aes(label = protein),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.color = "grey50"
  ) +
  scale_color_manual(values = c("Significant" = "red",
                                "Not significant" = "grey")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black") +
  scale_x_continuous(
    limits = c(-max(abs(results_df$log2FC), na.rm = TRUE),
                max(abs(results_df$log2FC), na.rm = TRUE))
  ) +
  theme_minimal() +
  labs(
    title = "Volcano plot: Olink proteins by sex",
    x = "log2 Median Difference (Male / Female)",
    y = "-log10 adjusted p-value",
    color = ""
  )

ggsave(
  filename = "volcano_olink_sex.png",
  plot = last_plot(),
  width = 8,
  height = 6,
  dpi = 300,
  bg = "white"
)

```





```{r}
olink_cols <- colnames(dat_de_master)[63:139]

sex_counts <- dat_de_master%>%
  filter(
    !is.na(dem_sex),
    rowSums(!is.na(select(., all_of(olink_cols)))) > 0
  ) %>%
  count(dem_sex)

sex_counts

```


Correlations 
```{r}
# Load packages
library(GGally)
library(ggplot2)
library(dplyr)

# Prepare data
pair_data <- dat_de_master %>%
  select(
    dem_sex,
    creatinine = Kreatinine1,
    BMI = anth_bmi,
    Age = dem_age
  ) %>%
  filter(complete.cases(.))

# Define colors for sex
sex_colors <- c(
  "Male"   = "#1f78b4",  # blue
  "Female" = "#fc9272"   # red
)

# Lower triangle: scatter + loess smooth
lower_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(aes(color = dem_sex), alpha = 0.6, size = 1.5) +
    geom_smooth(aes(color = dem_sex), method = "loess", se = FALSE) +
    scale_color_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Diagonal: density plots
diag_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_density(aes(fill = dem_sex), alpha = 0.5) +
    scale_fill_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Create the ggpairs plot and assign to an object
ggpairs(
  pair_data,
  columns = 2:4,   # exclude dem_sex
  lower = list(continuous = lower_fun),
  diag  = list(continuous = diag_fun))


```

Correlations_lactate
```{r}
# Load packages
library(GGally)
library(ggplot2)
library(dplyr)

# Prepare data
pair_data <- dat_de_master_lipid %>%
  select(
    dem_sex,
    Lactate = Lactate,
    BMI = anth_bmi,
    Age = dem_age
  ) %>%
  filter(complete.cases(.))

# Define colors for sex
sex_colors <- c(
  "Male"   = "#1f78b4",  # blue
  "Female" = "#fc9272"   # red
)

# Lower triangle: scatter + loess smooth
lower_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(aes(color = dem_sex), alpha = 0.6, size = 1.5) +
    geom_smooth(aes(color = dem_sex), method = "loess", se = FALSE) +
    scale_color_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Diagonal: density plots
diag_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_density(aes(fill = dem_sex), alpha = 0.5) +
    scale_fill_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Create the ggpairs plot and assign to an object
ggpairs(
  pair_data,
  columns = 2:4,   # exclude dem_sex
  lower = list(continuous = lower_fun),
  diag  = list(continuous = diag_fun))



```
DM duration
```{r}
# Load packages
library(GGally)
library(ggplot2)
library(dplyr)

# Prepare data
pair_data <- dat_de_master_lipid %>%
  select(
    dem_sex,
    Lactate = Lactate,
    BMI = anth_bmi,
    "DM Duration" = dm_dur_in_years,
  ) %>%
  filter(complete.cases(.))

# Define colors for sex
sex_colors <- c(
  "Male"   = "#1f78b4",  # blue
  "Female" = "#fc9272"   # red
)

# Lower triangle: scatter + loess smooth
lower_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(aes(color = dem_sex), alpha = 0.6, size = 1.5) +
    geom_smooth(aes(color = dem_sex), method = "loess", se = FALSE) +
    scale_color_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Diagonal: density plots
diag_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_density(aes(fill = dem_sex), alpha = 0.5) +
    scale_fill_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Create the ggpairs plot and assign to an object
ggpairs(
  pair_data,
  columns = 2:4,   # exclude dem_sex
  lower = list(continuous = lower_fun),
  diag  = list(continuous = diag_fun))



```
DM duration and creatininne


DM duration
```{r}
# Load packages
library(GGally)
library(ggplot2)
library(dplyr)

# Prepare data
pair_data <- dat_de_master_lipid %>%
  select(
    dem_sex,
    Creatinine = Kreatinine1,
    BMI = anth_bmi,
    "DM Duration" = dm_dur_in_years,
  ) %>%
  filter(complete.cases(.))

# Define colors for sex
sex_colors <- c(
  "Male"   = "#1f78b4",  # blue
  "Female" = "#fc9272"   # red
)

# Lower triangle: scatter + loess smooth
lower_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(aes(color = dem_sex), alpha = 0.6, size = 1.5) +
    geom_smooth(aes(color = dem_sex), method = "loess", se = FALSE) +
    scale_color_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Diagonal: density plots
diag_fun <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_density(aes(fill = dem_sex), alpha = 0.5) +
    scale_fill_manual(values = sex_colors) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Create the ggpairs plot and assign to an object
ggpairs(
  pair_data,
  columns = 2:4,   # exclude dem_sex
  lower = list(continuous = lower_fun),
  diag  = list(continuous = diag_fun))



```

