---
title: "PROTID_Olink"
author: "Lucy"
date: "2023-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1.1 Set directory and load libraries
```{r}
  ###data_dir <- "D:\Trainings\PhD\Data\Olink\Figures"
  output_dir <- "D:/Trainings/PhD/Data/Olink/Figures"
```


#1. Data pre-processing 
1.1 Loading data and subsetting PROTID IDs
```{r}
library("readxl")

PROTID_Olinkmain <-read_excel("D:/Trainings/PhD/Data/Proteomic_Analysis/Protid study_NPX_Clean.xlsx")### 271 Observations 95 variables 

## Remove controls from Olink and 300FG; leave in LOD, QCwarning and Plate id
# Load the "dplyr" package for data manipulation
library(dplyr)

# Filter for rows with specific "PID" values and include additional columns
PROTID_Olink <- PROTID_Olinkmain %>%
  filter(grepl("^13|^14|LOD", PID)) %>%
  select(PID, Plate_ID, QC_Warning, everything())

# Print the updated dataset
print(PROTID_Olink)
```



#2. Quality control
2.1 Calculate of % of missing values per protein in the dataset
```{r}
library(dplyr)

# Calculate the percentage of NA values in the "IL8" column
na_percentage <- sum(is.na(PROTID_Olink$IL8)) / nrow(PROTID_Olink) * 100

# Print the result
print(na_percentage)

# Calculate the percentage of NA values for all columns
na_percentage <- PROTID_Olink %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "NA_{.col}"))

# Print the result
print(na_percentage)
```


In long format
```{r}
# Load the required packages
library(dplyr)
library(broom)

# Calculate the percentage of NA values for all columns
na_percentage <- PROTID_Olink %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100, .names = "NA_{.col}")) %>%
  tidy()

# Print the result
print(na_percentage)
```
Which rows have NA?
```{r}
# Load the "dplyr" package for data manipulation
library(dplyr)

# Identify rows with NA values in any column
na_rows <- PROTID_Olink %>%
  filter(rowSums(is.na(.)) > 0)

# Print the result
print(na_rows)
```

Drop the above IDs that have missing values for all the proteins
```{r}
# Drop rows with specific "PID" values
PROTID_Olink <- PROTID_Olink %>%
  filter(!(PID %in% c("130118", "140119")))

# Print the updated dataset
print(PROTID_Olink)### 242 IDS
```

2.2 Calculate batch effect and stratify by site 13 Moshi 14 Mbeya
```{r}
## 5.1 Detect batch effect between plates

 PROTID_Olink$Plate_ID <- as.factor(PROTID_Olink$Plate_ID)
library(ggbiplot)
library(RColorBrewer)

  pca_df <- PROTID_Olink[-242,c(2,4:95)]
  prin_comp <- prcomp(pca_df[,-1], center = T,scale. = T)
  
  mycolor = colorRampPalette(brewer.pal(4, "Paired"))(length(unique( pca_df$Plate_ID)))
  ggbiplot(prin_comp, obs.scale = 1, var.scale = 1,
                groups = as.factor( pca_df$Plate_ID),
                circle = F, ellipse = T, var.axes = F)+
    scale_color_manual(values = mycolor,
                        name="Plates")+
    theme(legend.direction = 'horizontal',
          legend.position = 'top')+
    theme_bw(base_size=10)
```
  
####By site

```{r}
# Create new column "Site"
PROTID_Olink$Site <- ifelse(substr(PROTID_Olink$PID, 1, 2) == "13", "Moshi",
                            ifelse(substr(PROTID_Olink$PID, 1, 2) == "14", "Mbeya", NA))
print(PROTID_Olink$Site)

PROTID_Olink$Site <- as.factor(PROTID_Olink$Site)
library(ggbiplot)
library(RColorBrewer)

  pca_df <- PROTID_Olink[-242, c(96, 4:95)]
prin_comp <- prcomp(pca_df[, -1], center = TRUE, scale. = TRUE)

mycolor <- colorRampPalette(brewer.pal(4, "Paired"))(length(unique(pca_df$Site)))
ggbiplot(prin_comp, obs.scale = 1, var.scale = 1,
         groups = as.factor(pca_df$Site),
         circle = FALSE, ellipse = TRUE, var.axes = FALSE) +
  scale_color_manual(values = mycolor, name = "Site") +
  theme(legend.direction = "horizontal", legend.position = "top") +
  theme_bw(base_size = 10)

```
2.21
```{r}

```






2.3 Remove plate column, Site
```{r}
PROTID_Olink <- PROTID_Olink %>% select(!Plate_ID)
PROTID_Olink <- PROTID_Olink %>% select(!Site)
```


2.4 Protein data with high % value below limit of detection
```{r, fig.width=15, fig.height=10}
# Calculate percentage value below lod
  percent_lod_count <- data.frame(matrix(NA,nrow=92,ncol = 1))
  colnames(percent_lod_count) <- "Percentage_below_lod"
  percent_lod_count$Proteins <- colnames(PROTID_Olink)[3:94]
      
  df <- PROTID_Olink[-242,c(3:94)]
  lod <- PROTID_Olink[242,c(3:94)]
      
  
for (i in 1:ncol(df)) {
  for (j in 1:ncol(lod)) {
    if (colnames(df)[i] == colnames(lod)[j]) {
      percent_lod_count[j, 1] <- round((sum(df[[i]] < lod[[j]], na.rm = TRUE) / sum(!is.na(df[[i]]))) * 100, digits = 1)
    }
  }
}

    
#Order the data based on % value below LOD    
  df_qc_1 <- percent_lod_count[,c(2,1)] %>% 
             arrange(desc(Percentage_below_lod))

  # Visualisation
library(ggpubr)
library(forcats)
    
  theme_set(theme_pubclean())
  df_qc_1 %>% mutate(Proteins = fct_reorder(Proteins,Percentage_below_lod), .desc=F) %>%
        ggplot(aes(Proteins, Percentage_below_lod)) +
        geom_linerange(
          aes(x = Proteins, ymin = 0, ymax = Percentage_below_lod),
          color = "blue4", size = 1.5,
          position = position_dodge(0.3)
        )+
        geom_text(
          aes(label = ifelse(Percentage_below_lod >= 25.0,
                             paste0(Percentage_below_lod,"%"),"")),
          position = position_dodge(0.8),
          vjust = -0.3, size = 2.5
        )+
        xlab("Olink inflammatory panel") + 
        ylab("Percentage of lower detection limit measurements")+ 
        theme(axis.text.x=element_text(angle=90, hjust=1, size = 4))
  
```


2.5 Filter protein data with >= 25% % value below limit of detection
```{r}

# Filter protein data with >= 25% % value below limit of detection
  exclude_proteins <- df_qc_1$Proteins[df_qc_1$Percentage_below_lod>=25]
  exclude_proteins #15 Proteins
  
 PROTID_Olink <- PROTID_Olink %>% 
                  select(!all_of(exclude_proteins)) %>%
                  filter(!PID == "LOD")#### 241 observations 77 Proteins 
```



2.6 QC warning and outlier detection
```{r}
# QC warning
  rownames(PROTID_Olink) <- PROTID_Olink$PID
  prin_comp <- prcomp(PROTID_Olink[,-c(1,2)], center = T,scale. = T)

  ggbiplot(prin_comp, obs.scale = 1, var.scale = 1, 
                labels = PROTID_Olink$PID,
                groups = PROTID_Olink$QC_Warning, 
                ellipse = TRUE, circle = F, var.axes = F)+
        scale_color_manual(values=c("goldenrod3","blue4"),
                           name="Sample with QC warning") +
        theme(legend.direction = 'horizontal', 
                     legend.position = 'top')+
        theme_bw(base_size=10)
```


2.7 Outlier detection
```{r}
# Outliers detection
# Based on certain threshold (3 SD from mean of PC1 and PC2)
  
  Outlier_check_combined <- prin_comp$x
  out <- apply(Outlier_check_combined, 2, function(x) which(abs(x - mean(x)) > 
                                                              (3 * sd(x)) ))[1:2]
  outliers <- unique(sub("^....", "", names(unlist(out, use.names=T))))
  
  # Draw another PCA to visualize the outliers
  df_pca <- PROTID_Olink %>%
            mutate(Outliers=ifelse(PID %in% outliers,"Yes","No"))
  
  prin_comp <- prcomp(df_pca[,-c(1,2)], center = T,scale. = T)

  ggbiplot(prin_comp, obs.scale = 1, var.scale = 1, 
                labels = df_pca$PID,
                groups = df_pca$Outliers, 
                ellipse = F, circle = F, var.axes = F)+
        scale_color_manual(values=c("goldenrod3","blue4"),name="Outliers") +
        theme(legend.direction = 'horizontal', 
                     legend.position = 'top')+
        theme_bw(base_size=10)
```


Remove the QC waring column
```{r}
PROTID_Olink <- PROTID_Olink %>% select(!QC_Warning)#78 variables
```



2.8 Distribution of the proteomics data
```{r, fig.height=25, fig.width=25}
##Histogram and correlation of proteomic data

library(reshape2)
library(corrplot)
library(ggplot2)

# Histogram plot
  df_hist <- melt(PROTID_Olink[,c(2:78)])
  ggplot(df_hist, aes(value)) +
    geom_density(alpha=.5, fill="goldenrod3") +
    facet_wrap(~variable, scales = 'free', nrow = 10)+
    theme_minimal()
  
ggsave(paste0(output_dir,"Normalised.pdf")
  , plot = last_plot(), device = NULL, path = NULL,
    scale = 1, width = 25, height = 25, units = 'in', dpi = 300)
```

2.9 correlation plot
````{r, fig.height=12, fig.width=12}
# Make a correlation plot
library(corrplot)
#install.packages("psych")
library(psych)

  cor_df <- PROTID_Olink[,-1]
  df <-  psych::corr.test(cor_df, cor_df, adjust = "fdr", method = "pearson")
  df_p <- df$p
  df_r <- df$r

  #plot.new()
  col <- colorRampPalette(c("blue4", "white", "red4"))
  pdf(paste0(output_dir, 'Correlation.pdf'), width = 12, height = 12)
  corrplot(as.matrix(df_r), p.mat=as.matrix(df_p), method = "color",
           sig.level = c(.0001, .001, .05), pch.cex = .9, col=col(100),
           insig = "label_sig", pch.col = "black", tl.col = "black", cex.main = 1,
           mar = c(1,0,3,1), main="", order = "hclust", hclust.method= "complete",
           addrect = 4)
  
    pdf(paste0(output_dir,'Correlation.pdf'),
    width = 12, height =12)
```



PCA of all the proteins without the clinical variables. 
```{r}
 prin_comp <- prcomp(PROTID_Olink[,c(2:93)], center = T, scale. = T)

  ggbiplot(prin_comp, obs.scale = 1, var.scale = 1, 
                ellipse = F, circle = F, var.axes = F)+
        theme(legend.direction = 'horizontal', 
              legend.position = 'top')+
        theme_bw(base_size=10)

```

###save the rds file 
```{r}
##saveRDS(PROTID_Olink, "PROTID_Olink2.rds")
POlink_clean <- readRDS("D:/Trainings/PhD/Data/Proteomic_Analysis/PROTID_Olink2.rds")
```



Merging with the clinical variables...........................
```{r}
# Merge proteomic data with metadata (total 311 participants)

 Metadata_clean241 <- readRDS("D:/Trainings/PhD/Analysis/Metadata_clean241.rds")
 PROTID_Olink2 <- readRDS("PROTID_Olink2.rds")

  dat_de_master <- merge(Metadata_clean241,PROTID_Olink2,by.0="PID") 
  dim(dat_de_master)
  
  saveRDS(dat_de_master, "Merged.rds")

```

